The zebrafish mutant dreammist implicates sodium homeostasis in sleep regulation

Version 1

Abstract

Abstract

Sleep is a nearly universal feature of animal behaviour, yet many of the molecular, genetic, and neuronal substrates that orchestrate sleep/wake transitions lie undiscovered. Employing a viral insertion sleep screen in larval zebrafish, we identified a novel gene, dreammist (dmist), whose loss results in behavioural hyperactivity and reduced sleep at night. The neuronally expressed dmist gene is conserved across vertebrates and encodes a small single-pass transmembrane protein that is structurally similar to the Na+,K+-ATPase regulator, FXYD1/Phospholemman. Disruption of either fxyd1 or atp1a3a, a Na+,K+-ATPase alpha-3 subunit associated with several heritable movement disorders in humans, led to decreased night-time sleep. Since atpa1a3a and dmist mutants have elevated intracellular Na+ levels and non-additive effects on sleep amount at night, we propose that Dmist-dependent enhancement of Na+ pump function modulates neuronal excitability to maintain normal sleep behaviour.

Introduction

The ability of animals to switch between behaviourally alert and quiescent states is conserved across the animal kingdom (Cirelli, 2009; Joiner, 2016). Fundamental processes that govern the regulation of sleep-like states are shared across species, such as the roles of circadian and homeostatic cues in regulating the time and amount of sleep, stereotyped postures, heightened arousal thresholds, and the rapid reversibility to a more alert state (Joiner, 2016). The near ubiquity of sleep implies that it serves ancient functions and is subject to conserved regulatory processes. However, many key molecular components that modulate sleep and wake states remain undiscovered.

Over the past two decades, investigations into sleep and arousal states of genetically tractable model organisms, such as Drosophila melanogaster, C. elegans, and Danio rerio (zebrafish) have uncovered novel molecular and neuronal components of sleep regulation through gain-and loss-of-function genetic screens (reviewed in Barlow and Rihel, 2017; Sehgal and Mignot, 2011). The power of screening approaches is perhaps best exemplified by the first forward genetic sleep screen, which identified the potassium channel shaker as a critical sleep regulator in Drosophila (Cirelli et al., 2005). This result continues to have a lasting impact on the field, as not only did subsequent sleep screening efforts uncover the novel Shaker regulator sleepless, (Koh et al., 2009), but investigations into Shaker’s beta subunit Hyperkinetic ultimately revealed a critical role for this redox sensor linking metabolic function to sleep (Bushey et al., 2007; Kempf et al., 2019).

Disparate screening strategies across model organisms continue to unveil novel sleep modulators in both invertebrate and vertebrate model systems. For example, the roles of RFamide receptor DMSR-1 in stress-induced sleep in C. elegans (Iannacone et al., 2017) and SIK3 kinase in modulating sleep homeostasis in mice (Funato et al., 2016) were identified in genetic screens. Moreover, a gain of function screening strategy in Drosophila revealed the novel sleep and immune regulator, nemuri (Toda et al., 2019), and a zebrafish overexpression screen uncovered the secreted neuropeptides neuromedin U and neuropeptide Y, which decrease and increase sleep, respectively (Chiu et al., 2016; Singh et al., 2017). The success of screening strategies in revealing novel sleep-wake regulatory genes suggests that more sleep signals likely remain to be discovered.

One of the lessons from these genetic screens is that many of the uncovered genes play conserved roles across species. For example, Shaker also regulates mammalian sleep (Douglas et al., 2007) and RFamides induce sleep in worms, flies, and vertebrates (Lee et al., 2017; Lenz et al., 2015). Nevertheless, not every invertebrate sleep-regulatory gene has a clear vertebrate homolog, while some known human sleep/wake regulators, such as the narcolepsy-associated neuropeptide hypocretin/orexin (Chemelli et al., 1999; Lin et al., 1999; Peyron et al., 2000; Sakurai, 2013), lack invertebrate orthologs. Therefore, genetic sleep screens in vertebrates are likely to provide added value in uncovering additional regulatory components required to control the initiation and amount of sleep in humans.

While sleep screening in mammals is feasible (Funato et al., 2016), it remains an expensive and technically challenging endeavour. With its genetic tractability, availability of high-throughput sleep assays (Rihel and Schier, 2013), and conserved sleep genetics, such as the hypocretin, melatonin, locus coeruleus, and raphe systems (Gandhi et al., 2015; Singh et al., 2015; Oikonomou et al., 2019; Prober et al., 2006), the larval zebrafish is an attractive vertebrate system for sleep screens. We took advantage of a collection of zebrafish lines that harbour viral-insertions in >3500 genes (Varshney et al., 2013) to perform a targeted genetic screen. We identified a short-sleeping mutant, dreammist, with a disrupted novel, highly conserved vertebrate gene that encodes a small single pass transmembrane protein. Sequence and structural homology to the Na+/K+ pump regulator FXYD1/Phospholemman suggests that Dreammist is a neuronal-expressed member of a class of sodium pump modulators that is important for regulating sleep-wake behaviour.

Results

Reverse genetic screen identifies dreammist, a mutant with decreased sleep

We used the ‘Zenemark’ viral-insertion based zebrafish gene knock-out resource (Varshney et al., 2013) to perform a reverse genetic screen to identify novel vertebrate sleep genes. This screening strategy offers several advantages compared to traditional chemical mutagenesis-based forward genetic screening approaches. First, unlike chemical mutagenesis, which introduces mutations randomly, viral insertions tend to target the 5’ end of genes, typically causing genetic loss of function (Sivasubbu et al., 2007). Second, because the virus sequence is known, it is straightforward to map and identify the causative gene in mutant animals. Finally, since viral insertions in the Zenemark collection are already mapped and sequenced, animals harbouring insertions within specific gene classes can be selected for testing (Figure S1A). This allowed us to prioritise screening of genes encoding protein classes that are often linked to behaviour, such as G-protein coupled receptors, neuropeptide ligands, ion channels, and transporters (Supplemental Data 1).

For screening, we identified zebrafish sperm samples from the Zenemark collection (Varshney et al., 2013) that harboured viral insertions in genes of interest and used these samples for in vitro fertilization and the establishment of F2 families. For each viral insertion line, clutches from heterozygous F2 in-crosses were raised to 5 days post-fertilisation (dpf) and tracked using videography (Figure S1A) to quantify the number and duration of sleep bouts (defined in zebrafish larvae as inactivity lasting 1 minute or longer; Prober et al., 2006) and waking activity (time spent moving per active bout) over 48 hours. The genotypes of individual larvae were determined by PCR after behavioural tracking, with each larva assigned as wild type, heterozygous, or homozygous for a given viral insertion to assess the effect of genotype on sleep/wake behaviour. While most screened lines had minimal effects on sleep-wake behavioural parameters (Figure S1B-S1C), one homozygous viral insertion line, 10543/10543, had a reduction in daytime sleep (Figure S1B) and an increase in daytime waking activity (Figure S1C) relative to their wild type sibling controls. We re-named this 10543 viral insertion line dreammist (dmist).

In follow-up studies, we observed that animals homozygous for the viral insertion at this locus (dmistvir/vir) showed a prominent decrease in sleep during the day and a trend to sleep less at night compared to their wild-type siblings (dmist+/+) (Figure 1A). dmist mutants had an almost 50% reduction in the average amount of daytime sleep (Figure 1C) due to a decrease in the number of sleep bouts (Figure 1D), whereas the sleep bout length at night was significantly reduced (Figure 1E). dmistvir/vir larvae also exhibited significantly increased daytime waking activity, which is the locomotor activity while awake (Figure 1B, 1F). Because Zenemark lines can contain more than one viral insertion (17.6% of lines have ≥2 insertions; Varshney et al 2013), we outcrossed dmistvir/+ fish to wild-type fish of the AB-TL background and re-tested dmist mutant fish over several generations. Normalising all the behavioural parameters to dmist+/+ controls with a linear mixed effects (LME) model showed consistent sleep changes in dmistvir/vir fish over 5 independent experiments (Figure 1G). The dmistvir/vir larvae consistently show a more than 50% decrease in sleep during the day due to a significant reduction in the number and duration of sleep bouts, as well as a large increase in waking activity (Figure 1G). The dmistvir/vir mutants also had a significant reduction in sleep at night compared to wild type siblings (Figure 1G). These effects on sleep and wakefulness are not due to alterations in circadian rhythms, as behavioural period length in fish that were entrained and then shifted to free-running constant dark conditions was unaffected in dmistvir/vir compared to wild-type sibling larvae (Figure S2A-S2C).

A viral insertion mini-screen identifies a short-sleeping mutant, dreammist.

A-B) Mean± SEM sleep (A) and waking activity (B) of progeny from dmistvir/+ in-cross from original screen. White blocks show day (lights on) and grey blocks show night (lights off). Data is combined from 2 independent experiments. n indicates the number of animals.

C-F) Analysis of sleep/wake architecture for the data shown in (A, B). C) Quantification of total sleep across two days and nights shows decreased day and night sleep in dmistvir/vir. Analysis of sleep architecture reveals fewer sleep bouts during the day (D) and shorter sleep bouts at night (E) in dmistvir/vir compared with sibling controls. F) Daytime waking activity is also increased in dmistvir/vir. The black lines show the mean ± SEM, except in E, which labels the median ± SEM. *p<0.05, **p<0.01, ***p<0.001; ns p>0.05; one-way ANOVA, Tukey’s post hoc test.

G) Combining 5 independent experiments using a linear mixed effects model with genotype as a fixed effect and experiment as a random effect reveals dmistvir/vir larvae have decreased total sleep and changes to sleep architecture during both the day and night compared to dmist+/+ siblings. Plotted are the genotype effect sizes (95% confidence interval) for each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05, **p<0.01, ***p<0.001, ns p>0.05. n indicates the number of animals.

The dmist gene encodes a novel, small transmembrane protein

Having identified a sleep mutant, we next sought to investigate the target gene disrupted by the viral insertion. Line 10543 (dmistvir) was initially selected for screening due to a predicted disruption of a gene encoding a serotonin transporter (slc6a4b) on chromosome 5. However, mapping of the dmist viral insertion site by inverse-PCR and sequencing revealed that the virus was instead inserted into the intron of a small two-exon gene annotated in the Zv6 genome assembly as a long intergenic non-coding RNA (lincRNA; gene transcript ENSDART00000148146, gene name si:dkey234h16.7), which lies approximately 6 kilobases (kb) downstream of the slc6a4b gene in zebrafish. At least part of this region is syntenic across vertebrates, with a small two-exon gene identified adjacent to the genes ankrd13a and GIT in several vertebrates, including human and mouse (Figure 2A). Amplifying both 5’ and 3’ ends of zebrafish si:dkey234h16.7 and mouse E13.5 1500011B03-001 transcripts with Rapid Amplification of cDNA ends (RACE) confirmed the annotated zebrafish and mouse transcripts and identified two variants with 3’ untranslated regions (3’UTR) of different lengths in zebrafish (Figure S3B). To test whether the viral insertion in dmistvir/vir disrupts expression of si:dkey234h16.7 or neighbouring genes, we performed quantitative analysis of gene transcript levels in wild type and mutant dmist larvae by RT-qPCR. This revealed that the dmist viral insertion caused a more than 70% reduction in the expression of si:dkey234h16.7 while the expression of the most proximal 5’ or 3’ flanking genes, slc6a4b_Dr and ankrd13a_Dr, were unaffected (Figure 2B and S3A). Since this reduced expression is most consistent with si:dkey234h16.7 being the causal lesion of the dmist mutant sleep phenotype, we renamed this gene dreammist (dmist).

dmist encodes a conserved vertebrate single pass transmembrane protein.

A) dmist mutants harbour a viral insertion in the 1st intron of si:key-234h16.7. dmist is syntenic with Ankrd13 and GIT orthologs in mouse, human, and zebrafish.

B) RT-qPCR of dmist (red) show reduced expression of dmist and not the 5’ and 3’ flanking zebrafish genes, slc6a4b (cyan) and ankrd13a (blue), in dmistvir/vir larvae compared to dmistvir/+ and dmist+/+ siblings. **p<0.01, *p<0.05; ns p>0.05; one-way ANOVA, Tukey’s post-hoc test. Data shows mean ± SEM normalized to the wild type mean.

C) dmist_Dr contains an open reading frame encoding a 70 amino acid protein that is conserved across vertebrates. All identified homologues have a predicted signal peptide sequence (magenta line), signal peptide cleavage site (magenta circle), and predicted transmembrane domain (grey), with additional highly conserved C-terminal motifs (blue lines). Identical amino acids in all species are shown in black; similar amino acids (80-99% conserved across species) are shown in grey.

D) In situ hybridisation using a dmist antisense probe reveals dmist is maternally deposited as it is detected at the 2-cell stage. At 24 hpf expression is restricted to regions containing neuronal precursors, and at 5 dpf expression is widespread throughout the brain. Tel, telencephalon; Dien, diencephalon; R1-6, rhombomeres 1-6; A, anterior; P, posterior. Scale bars= 0.5 mm (2 cell and 24 hpf), 0.1 mm (5 dpf).

E-F) Representative confocal image of 90% epiboly embryo co-injected at the 1-cell stage with mRNA encoding membrane-RFP (magenta) and a plasmid encoding either C-terminal tagged Dmist-GFP (E, green) or DmistA22W-GFP (F, green). Scale bar= 25 μm.

Computational predictions indicated that the dmist transcripts contain a small open reading frame (ORF) encoding a protein of 70 amino acids (aa) (Figure 2C, Figure S3B). Querying the human and vertebrate protein databases by BLASTp using the C-terminal protein sequence of Dmist identified orthologs in most vertebrate clades, including other species of teleost fish, birds, amphibians, and mammals (Figure 2A, C). All identified orthologs encoded predicted proteins with an N-terminal signal peptide sequence and a C-terminal transmembrane domain (Figure 2C). The peptide sequence identity across orthologs ranged from 38 to 84%, with three peptide motifs (QNLV, CVYKP, RRR) showing high conservation across all vertebrates, and high similarity for many additional residues (Figure 2C, Figure S3D). Additional searches by tBLASTn failed to identify any non-vertebrate dmist orthologs. In summary, we found that the dreammist gene, the expression of which is disrupted in dmistvir/vir fish with sleep phenotypes, encodes a protein of uncharacterized function that is highly conserved across vertebrates at both the genomic and molecular levels.

Genetic molecular analysis of dmist expression in zebrafish and mouse

Because the viral insertion disrupts dmist throughout the animal’s lifetime, we examined both the developmental and spatial expression of dmist to assess when and where its function may be required for normal sleep. Using the full-length transcript as a probe (Figure S3B), we performed in situ hybridization across embryonic and larval zebrafish development. Maternally deposited dmist was detected in early embryos (2-cell stage) prior to the maternal to zygotic transition (Giraldez et al., 2006) (Figure 2D). Consistent with maternal deposition of dmist transcripts, inspection of the 3’ end of the dmist gene revealed a cytoplasmic polyadenylation element (‘TTTTTTAT’; Supplemental Information 2) that is required for zygotic translation of maternal transcripts (Villalba et al., 2011). At 24 hpf, transcripts were detected in regions that form the embryonic brain, such as ventral telencephalon, diencephalon and cerebellum, and in the developing eye (Figure 2D, S3C). By 5 dpf, dmist transcripts were detected throughout the brain (Figure 2D). To test whether dmist transcripts are under circadian regulation, we performed RT-qPCR in fish that were entrained and then shifted to free-running constant dark conditions. In contrast with the robust 24-hr rhythmic transcription of the circadian clock gen per1, we did not detect any changes in dmist expression throughout the 24 hour circadian cycle (Figure S2D).

Consistent with brain expression in larval zebrafish, we identified the expression of Dmist_Mm in a published RNAseq dataset of six isolated cell types from mouse cortex (Zhang et al., 2014). We confirmed that Dmist_Mm is specifically enriched in neurons by hierarchical clustering of all 16,991 expressed transcripts across all six cells types, which demonstrated that Dmist_Mm co-clusters with neuronal genes (Figure S3E). Pearson correlation of Dmist_Mm with canonical markers for the six cell types showed that Dmist_Mm expression is highly correlated with other neuronal genes but not genes associated with microglia, oligodendrocytes, or endothelia. This result indicates that dmist is specifically expressed in neurons in both zebrafish and mouse (Figure S3F).

Dmist localises to the plasma membrane

Although the dmist gene encodes a conserved ORF with a predicted signal peptide sequence and transmembrane domain (Figure 2C; Figure S3G-I), we wanted to confirm this small peptide can localise to the membrane and if so, on which cellular compartments. To test these computational predictions, we transiently co-expressed GFP-tagged Dmist (C-terminal fusion) with a marker for the plasma membrane (myr-Cherry) in zebrafish embryos. Imaging at 90% epiboly revealed Dmist-GFP localised to the plasma membrane (Figure 2E). Conversely, introducing a point mutation into Dmist’s signal peptide cleavage site (DmistA22W-GFP) prevented Dmist from trafficking to the plasma membrane, with likely retention in the endoplasmic reticulum (Figure 2F). Together, these data indicate that Dmist localises to the plasma membrane despite its small size, as computationally predicted.

CRISPR/Cas9 generated dmisti8 mutant exhibits decreased night-time sleep

dmist expression was reduced by 70% in the viral insertion line, suggesting that dmistvir is a hypomorphic allele. To confirm that the sleep phenotypes observed in dmistvir/vir animals are due to the loss of Dmist function, we used CRISPR/Cas9 to create an independent dmist loss of function allele. We generated a zebrafish line in which the dmist gene contains an 8 bp insertion that causes a frameshift and early stop codon (dmisti8, Figure 3A). The dmisti8 allele is predicted to encode a truncated protein lacking the complete signal peptide sequence and transmembrane domain (Figure 3B), indicating this is likely a null allele. RT-qPCR showed that dmist transcript levels were 60% lower in dmisti8/i8 fish compared to wild type siblings, consistent with nonsense-mediated decay (Figure S4A, B).

CRISPR-generated dmist mutants sleep less and are hyperactive at night.

A) CRISPR/Cas9 targeting of the first exon of dmist resulted in an 8 bp insertion (dmisti8) (grey line) within the coding sequence, leading to an early stop codon (red line with *). Guide RNA target sequence and PAM sequence are shown as black bars. The sequence that is deleted in the mutant is indicated with a red bar.

B) Predicted Dmisti8 peptide sequence lacks most of the N-terminal signal peptide sequence (magenta) and the full C-terminus.

C-D) Representative 48 hr traces of mean ± SEM sleep (C) and waking activity (D) shows decreased sleep and increased waking activity at night for dmisti8/i8 fish compared to dmisti8/+ and dmist+/+ siblings. n=number of fish.

E-H) Analysis of sleep/wake architecture of the experiment depicted in (C, D) indicates that dmisti8/i8 larvae sleep less at night (E) due to fewer sleep bouts (F). Sleep bout length is unchanged (G). Waking activity is also increased in dmisti8/i8 fish (H). The black line represents the mean ± SEM except for G, which is the median ± SEM. *p<0.05, **p<0.01, ***p<0.001; One-way ANOVA, Tukey’s post hoc test.

I) Combining 5 independent experiments with a linear mixed effects model reveals dmisti8/i8 fish sleep less at night due to fewer sleep bouts and also show increased waking activity at night. Plotted are the genotype effect sizes (95% confidence interval) for each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05, **p<0.01, ***p<0.001, ns p>0.05.

J-K) (J) Adult dmisti8/i8 fish have a higher mean swim speed compared to their wild type siblings at night. Data in (J) is quantified at night in (K). (J, K) show mean ± SEM. *p<0.05, one-way ANOVA.

L) Cumulative probability distribution of all night-time swim bout speeds in adult fish. The dashed lines show the half max (0.5 probability) for each curve. *p<0.05 for dmisti8/i8 fish compared to wild type siblings; Kolmogorov-Smirnov test.

We next assessed the sleep and activity patterns of dmisti8/i8 fish. As seen in exemplar individual tracking experiments, dmisti8/i8 larvae sleep less at night due to fewer sleep bouts and also show an increase in waking activity relative to wild type and heterozygous mutant siblings (Figure 3C-H). This significant night-time reduction in sleep and increase in hyperactivity is also apparent when combining 5 independent experiments with an LME model to normalize behaviour across datasets (Figure 3I). Although dmistvir/vir larvae also sleep less at night (Figure 1G), the large day-time reduction in sleep observed in dmistvir/vir larvae is absent in dmisti8/i8 animals, perhaps due to differences in genetic background that affect behaviour. Because the dmistvir is likely a hypomorphic allele, we focused subsequent experiments on the CRISPR-generated dmisti8/i8 larvae.

To test whether the increased night-time activity of dmisti8/i8 mutants persists in older animals, we raised dmisti8/i8 mutants with their heterozygous and wild type siblings to adulthood in the same tank and tracked individual behaviour for several days on a 14:10 light:dark cycle. As in larval stages, dmisti8/i8 adults were hyperactive relative to both dmisti8/+ and dmist+/+ siblings, maintaining a higher mean speed at night (Figure 3J-L). This suggests that either Dmist affects a sleep/wake regulatory circuit during development that is permanently altered in dmist mutants, or that Dmist is continuously required to maintain normal levels of night-time locomotor activity.

Dmist is distantly related to the Na+/K+ pump regulator Fxyd1 (Phospholemman)

Because Dmist is a small, single pass transmembrane domain protein without any clear functional motifs that has not been functionally characterized in any species, we searched for similar peptides that might provide clues for how Dmist regulates behaviour. Using the multiple sequence alignment tool MAFFT to align the zebrafish, mouse, and human Dmist peptides (Katoh and Toh, 2010) and seeding a JackHMMR iterative search of the Uniprot database (Johnson et al., 2010) found distant homology between Dmist and Fxyd1/Phospholemman (Figure 4A), a small transmembrane domain peptide that regulates ion channels and pumps, including the Na+,K+-ATPase pump (Crambert et al., 2002). Dmist and Fxyd1 share 27-34% amino acid homology, including an RRR motif at the C-terminal end, although Dmist lacks a canonical FXYD sequence (Figure 4A). In addition, computational predictions using the AlphaFold protein structure database revealed structural similarities between Dmist and Fxyd1 (Jumper et al., 2021), suggesting that Dmist may belong to a class of small, single pass transmembrane ion pump regulators.

Mutation of the dmist related gene fxyd1 causes reduced sleep at night.

A) Schematic of zebrafish Dmist and Fxyd1 protein domains and alignments comparing human, mouse, and zebrafish Dmist and FXYD1 protein sequences. Black and grey shading indicate amino acid identity and similarity, respectively. The FXYD domain is indicated with a red line and the RRR motif in the C-terminus is indicated with a dark blue line.

B) CRISPR-Cas9 targeting of the 3rd exon of fxyd1 created a 28 bp deletion, resulting in a predicted truncated protein. The start codon is marked by a cyan line. Guide RNA target sequence and PAM sequence are shown as black bars. The mutant deleted sequence is indicated with a red bar.

C) In situ hybridisation of fxyd1 at 24 hpf (whole animal) and 5 dpf brain (ventral view). Anterior is to the left. Scale bar = 0.5 mm (24 hpf); 0.1 mm (5 dpf).

D-E) Representative behavioral experiment showing fxyd1Δ28 mutants have decreased night-time sleep (D) but normal waking activity at night (E).

F) Combining 5 independent experiments with a linear mixed effects model reveals fxyd1Δ28/Δ28 larvae sleep significantly less at night due to shorter sleep bouts compared to fxyd1+/+ siblings. Plotted are the genotype effect sizes (95% confidence interval) on each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05, **p<0.01, ***p<0.001, ns p>0.05.

Using In situ hybridisation, we found that fxyd1 is expressed in cells along the brain ventricle and choroid plexus (Figure 4C) in contrast to the neuronal expression of dmist (Figure 2D). Despite these different expression patterns, based on their sequence similarity we reasoned that Fxyd1 and Dmist may regulate the same molecular processes that are involved in sleep. To test this hypothesis, we used CRISPR/Cas9 to generate a 28 bp deletion in the third exon of the zebrafish fxyd1 gene, causing a frameshift that is predicted to encode a truncated protein that lacks the FXYD, transmembrane, and C-terminal domains (Figure 4B). Contrary to a previous report based on morpholino knockdown (Chang et al., 2012), fxyd1Δ28/Δ28 larvae were viable with no detectable defect in inflation of the brain ventricles. We therefore tested fxyd1 mutant larvae for sleep phenotypes. Like dmist mutants, fxyd1Δ28/Δ28 larvae slept less at night (Figure 4D-F). Interestingly, this sleep loss is mainly due to shorter sleep bouts (Figure 4F), indicating that fxyd1 mutants initiate sleep normally but do not properly maintain it, unlike dmist mutants, which initiate fewer night-time sleep bouts, although in both cases there is consolidation of the wake state at night (Figure 3I, 4F). Thus, despite the non-neuronal expression of fxyd1 in the brain, mutation of the gene most closely related to dmist results in a similar sleep phenotype.

The brain-wide Na+/K+ pump alpha subunit Atp1a3a regulates sleep at night

Given the similarity between Dmist and Fxyd1 and their effects on night-time sleep, we hypothesized that mutations in Na+/K+ pump subunits known to interact with Fxyd1 might also affect sleep. Consistent with this hypothesis, a low dose of the Na+/K+ pump inhibitor, ouabain, reduced night-time sleep in dose-response studies (Figure S5A). When applied in the late afternoon of 6 dpf, 1 µM ouabain decreased subsequent night-time sleep by 16.5% relative to controls, an effect size consistent with those observed in dmist mutants (Figure 6A, C). Night-time waking activity was also significantly increased after low-dose ouabain exposure (Figure 6B, D). Ouabain binds to specific sites within the first extracellular domain of Na+/K+ pump alpha subunits (Price and Lingrel, 1988), and species-specific changes to these sites confers species-specific ouabain resistance, as in the case of two naturally occurring amino acid substitutions present in the Atp1a1 subunit of mice (Dostanic et al., 2004). Alignment of the ouabain sensitive region of zebrafish and mouse Na+/K+ pump alpha subunits revealed that zebrafish Atp1a1a lacks the conserved Glutamine at position 121 (Figure 5E), suggesting that one of the other subunits with conserved ouabain-binding sites is responsible for the low dose ouabain sleep effects. We focused on the Na+/K+ pump alpha-3 subunit (Atp1a3), as this has been shown to directly interact with Fxyd1 in mammalian brain tissue (Feschenko et al., 2003). Murine Dmist expression also correlates well with the Atp1a3 distribution across 5 brain cell types in mouse (Pearson correlation coefficient = 0.63), which has the strongest correlation score with neuronal markers (Figure S5B compared to Figure S3F). In contrast, zebrafish atp1a2a is reportedly expressed in muscle at larval stages, while atp1a1b is confined to cells along the ventricle (Thisse et al., 2001).

Mutation of the Na+/K+ pump alpha subunit atp1a3a reduces sleep at night

A-B) Mean ± SEM sleep and waking activity traces of wild type larvae following exposure to 1 µM ouabain. Arrows indicate time the drug was added.

C-D) At night, sleep is significantly reduced and waking activity is significantly increased after ouabain exposure. Student’s t-test, one tailed.

E) Alignments of Na+/K+ pump alpha subunits around the ouabain binding sites. Red indicates residues that are critical for higher sensitivity to ouabain, which are present in mouse Atp1a3 but not Atp1a1.

F) In situ hybridisation of atp1a3a at 24hpf (whole animal) and 5dpf brain (ventral view). Anterior is to the left. Scale bar = 0.5 mm (24 hpf); 0.1 mm (5 dpf). A-anterior; P-posterior; D-Dorsal; V-Ventral

G) CRISPR-Cas9 targeting of the atp1a3a resulted in a 19 bp deletion that eliminates the start codon (blue) and splice junction. Guide RNA target sequence and PAM sequence are shown as black bars. Sequence that is deleted in the mutant is indicated with a red bar.

H-I) Representative behavioural experiment showing atp1a3aΔ19/Δ19 fish are hyperactive throughout the day-night cycle and have decreased sleep at night. Mean ± SEM are shown.

J) atp1a3aΔ19/Δ19 larvae sleep less at night due to shorter sleep bouts. Plotted are the genotype effect sizes (95% confidence interval) on each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05, **p<0.01, ***p<0.001, ns p>0.05.

dmist mutants have altered sodium homeostasis

A) Brain sodium levels are significantly elevated after exposure to PTZ in both atp1a3aΔ19/Δ19 (2 independent experiments) and dmisti8/i8 (4 independent experiments) fish relative to wild type and heterozygous mutant siblings, as measured by fluorescence intensity of Sodium Green, normalized to the sample mean intensity. Crosses show mean ± SEM. n indicated the number of animals. Below are example images of brains stained with Sodium Green. *p<0.05, **p<0.01, one-way ANOVA, Tukey’s post hoc test.

B) Under baseline conditions, brain sodium levels are significantly elevated in dmisti8/i8 fish at night but not during the day, as measured by fluorescence intensity with Sodium Green. Crosses show mean ± SEM. *p<0.05, **p<0.01, one-way ANOVA, Tukey’s post hoc test.

C) dmisti8/i8 larvae have increased rebound sleep compared to wild type siblings following exposure to 5mM PTZ. Representative sleep traces of dmist+/+ (no drug, water vehicle controls in black; PTZ exposed in blue) and dmisti8/i8 (no drug in purple; PTZ exposed in red) following 1 hr exposure to 5 mM PTZ (black bar) in the morning. Data are mean ± SEM. dmisti8/+ animals are not plotted for clarity but are included in panel D.

D) Rebound sleep after exposure to 5 mM PTZ, calculated from the experiment in C. Each dot represents a single fish, grey lines show mean ± SEM.

E) Effect size of change in sleep after 1 hr treatment with 5 mM PTZ (and washout) compared to vehicle treated controls (error bars show 95% confidence intervals). *p<0.05, one-way ANOVA, Tukey’s post-hoc test.

F) Effect sizes (and 95% confidence interval) relative to wild types (dotted line) on sleep at night in larvae from dmist+/-; atp1a3a+/- in-crosses from 3 independent experiments. P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model relative to dmist+/+; atp1a3a+/+ animals. For all sleep-wake parameters, see Figure S6. *p<0.05, **p<0.01, ***p<0.0001, ns p>0.05.

Zebrafish have two Atp1a3 paralogs, atp1a3a and atp1a3b. Similar to dmist, atp1a3a is widely expressed in the larval zebrafish brain (Figure 5F, compare to Figure 2D). While atp1a3b is also expressed in the zebrafish brain, its expression is more limited to regions of the midbrain and hindbrain (Figure S5C). To test whether these genes are involved in regulating zebrafish sleep, we used CRISPR/Cas9 to isolate an allele of atp1a3a containing a 19 bp deletion and an allele of atp1a3b containing a 14 bp deletion. Both mutations are predicted to generate null alleles due to deletion of the start codon (Figure 5G, S5D). Both atp1a3a Δ19/Δ19 and atp1a3b Δ14/Δ14 mutant larvae were healthy and viable through early development, although atp1a3b mutant larvae were not obtained at Mendelian ratios. Contrary to a previous report based on morpholino injections (Doǧanli et al., 2013), neither mutant had defects in the inflation of their brain ventricles. Sleep-wake tracking experiments found that atp1a3b Δ14/Δ14 mutants were more active during the day with minimal sleep phenotypes (Figure S5E-G). In contrast, mutation of atp1a3a resulted in large effects on sleep-wake behaviour. Compared to wild type and heterozygous mutant siblings, atp1a3aΔ19/Δ19 animals were hyperactive throughout the day and night and had a large reduction in sleep at night (Figure 5H, I). The night-time sleep reduction was due to a reduction in the length of sleep bouts, as atp1a3a mutants even had a small increase in the number of sleep bouts at night (Figure 5J). In conclusion, loss of atp1a3a results in sleep loss at night, similar to treatment with the small molecule ouabain, and to dmist and fxyd1 mutants. Notably, the atp1a3a mutant phenotype is much stronger, as might be expected if Dmist plays a modulatory, and Atp1a3a a more central, role in Na+/K+ pump activity.

Dmist modulates Na+/K+ pump function and neuronal activity-induced sleep homeostasis

The similar night-time reduction in sleep in dmist and atp1a3a mutants, combined with the similarities between Dmist and Fxyd1, suggested that Dmist may regulate the Na+/K+ pump. We therefore exposed wild type and mutant larvae to pentylenetetrazol (PTZ), a GABA-receptor antagonist that leads to globally heightened neuronal activity and elevated intracellular sodium levels that must be renormalized by Na+/K+ pump activity. Consistent with the hypothesis that Dmist and Atp1a3a subunits are important for a fully functional Na+/K+ pump, brains from both dmisti8/i8 and atp1a3aΔ19/Δ19 larvae had elevated intracellular sodium levels after exposure to PTZ (Figure 6A). Thus, neither dmist nor atp1a3a mutants were able to restore intracellular sodium balance after sustained neuronal activity as quickly as wild type siblings. Consistent with the night-specific alterations in sleep behaviour, we also found that baseline brain Na+ levels in dmist mutants were significantly elevated at night but not during the day (Figure 6B). Collectively, these data are consistent with the hypothesis that night-time sleep duration is affected by changes in Na+/K+ pump function and that Dmist is required to maintain this function both at night and after sustained high levels of neuronal activity.

We have previously shown in zebrafish that a brief exposure to hyperactivity-inducing drugs such as the epileptogenic PTZ or wake-promoting caffeine induces a dose-dependent increase in homeostatic rebound sleep following drug washout that is phenotypically and mechanistically similar to rebound sleep following physical sleep deprivation (Reichert et al., 2019). Based on their exaggerated intracellular Na+ levels following exposure to PTZ, we predicted that dmist mutants would also have increased rebound sleep in response to heightened neuronal activity. Upon wash-on/wash-off of lower dose (5 mM) PTZ, sleep rebound occurs in approximately 50% of wild type larvae (Reichert et al., 2019; Figure 6C, D). In contrast, all dmisti8/i8 larvae showed increased rebound sleep compared to dmist+/+ sibling controls (Figure 6C-E). Taken together with the elevated sodium retention experiments, such increases in rebound sleep induced by neuronal activity suggests that dmisti8/i8 fish more rapidly accumulate sleep pressure in response to heightened neuronal activity.

Finally, we predicted that if Dmist is affecting baseline sleep via modulation of Atp1a3a-containing Na+/K+ pumps, dmist-/-; atp1a3a-/- double mutants should have a reduction in night-time sleep that is not the sum of effects from either mutant alone. In other words, if Dmist and Atp1a3a are acting in separate pathways, the double mutant would have an additive phenotype, but if Dmist and Atp1a3a act together in the same complex/pathway, the mutant phenotypes should be non-additive. Indeed, dmist-/-; atp1a3a-/- mutants have a sleep reduction similar to that of atp1a3a-/- mutants alone, consistent with a non-additive effect (Figure 6F and S6). Similar non-additivity can be also observed in the dmist-/-; atp1a3a+/- animals, which, like atp1a3a+/- animals alone, have a milder sleep reduction, indicating that the lack of additivity between dmist and atp1a3a phenotypes is unlikely due to a floor effect, since double homozygous mutants can sleep even less (Figure 6F). This genetic interaction data is consistent with our hypothesis that Atp1a3a and Dmist act in the same pathway—the Na+/K+ pump--to influence sleep.

Discussion

Genetic screening discovers dmist, a novel sleep-regulatory gene

Using a reverse genetic viral screening strategy, we discovered a short-sleeping mutant, dmist, which has a disruption in a previously uncharacterized gene encoding a small transmembrane peptide. Given that the dmist mutant appeared within the limited number of 25 lines that we screened, it is likely that many other sleep genes are still waiting to be discovered in future screens. In zebrafish, one promising screening strategy will be to employ CRISPR/Cas9 genome editing to systematically target candidate genes. Advances in the efficiency of this technology now makes it feasible to perform a CRISPR “F0 screen” in which the consequences of bi-allelic, gene-specific mutations are rapidly tested in the first generation, with only the most promising lines pursued in germline-transmitted mutant lines (Grunwald et al., 2019; Jao et al., 2013; Kroll et al., 2021; Shah et al., 2015; Shankaran et al., 2017; Wu et al., 2018). CRISPR F0 screens could be scaled to systematically target the large number of candidate sleep-regulatory genes identified through human GWAS studies and sequencing of human patients suffering from insomnia and neuropsychiatric disorders (Allebrandt et al., 2013; Dashti et al., 2019; Jansen et al., 2019; Jones et al., 2019; Lane et al., 2019; Lek et al., 2016; Palagini et al., 2019).

Dmist is related to the Na+/K+ pump regulator Fxyd1

The small Dmist transmembrane protein is highly conserved across vertebrates, expressed in neurons, and important for maintaining normal sleep levels. How can such a small, single pass transmembrane protein lacking any clear functional domains modulate the function of neurons and ultimately animal behaviour? The recognition that Dmist has sequence homology (∼35% amino acid similarity; a conserved ‘RRR’ motif in the C-terminus) and structural homology (e.g. signal peptide and single pass transmembrane domains) to the Na+,K+-ATPase pump regulator Fxyd1 (Phospholemman) offers some important clues.

Fxyd1/Phospholemman is a member of the FXYD protein family, of which there are seven mammalian members (Sweadner and Rael, 2000). Each of the FXYD proteins is small, contains a characteristic FXYD domain, and has a single transmembrane domain. FXYD family members interact with alpha subunits of the Na+,K+ ATPase to regulate the function of this pump, with individual family members expressed in different tissues to modulate Na+,K+-ATPase activity depending on the physiological needs of the tissue (Geering et al., 2003). In cardiac muscle, FXYD1 is thought to act as a hub through which various signalling cascades, such as PKA, PKC, or nitric oxide, can activate or inhibit Na+ pump activity (Pavlovic et al., 2013). For example, FXYD1 is critical for mediating the increased Na+ pump activity observed after β-receptor stimulation via cAMP-PKA signalling (Despa et al., 2008). Much less is known about the role of FXYD1 in non-cardiac tissue, although it is expressed in neurons in the mammalian cerebellum, the choroid plexus, and ependymal cells, where it interacts with all three alpha subunits of the Na+,K+ ATPase (Feschenko et al., 2003).

In zebrafish, we also found that fxyd1 is expressed in cells around the ventricles and in the choroid plexus (Figure 4C), in contrast to dmist which is expressed in neurons throughout the brain. Despite the different expression patterns, mutation of each gene resulted in a similar reduction of sleep at night. However, unlike dmist mutants, which have fewer sleep bouts (i.e. initiate sleep less) and an increase in waking locomotor activity, fxyd1 mutants have shorter sleep bouts (i.e. cannot maintain sleep) on average and do not have a locomotor activity phenotype. Just as the various FXYD family members modulate the Na+/K+ pump in different tissue-and context-specific ways, this phenotypic variation between fxyd1 and dmist mutants could be due to the different fxyd1 and dmist expression patterns, modulation kinetics of pump/channel dynamics, or interaction with different accessory proteins or signal transduction cascades. Nevertheless, the similar timing and magnitude of sleep reduction, combined with the structural similarity of Fxyd1 and Dmist, suggest that they may regulate similar sleep-related processes.

Dmist, the sodium pump, and sleep

The similarity between Dmist and FXYD1 led us to directly manipulate the Na+,K+ ATPase to test its importance in sleep. The Na+,K+-ATPase is the major regulator of intracellular Na+ in all cells and, by actively exchanging two imported K+ ions for three exported Na+ ions, is essential for determining cellular resting membrane potential (reviewed in Clausen et al., 2017). The Na+,K+-ATPase consists of a catalytic alpha subunit (4 known isoforms, ATP1A1-4), a supporting beta subunit (3 isoforms, ATP1B1-3), and a regulatory gamma subunit (the FXYD proteins). The alpha1 and alpha3 subunits are the predominant catalytic subunits in neurons (alpha2 is mostly restricted to glia), although the alpha1 subunit is also used ubiquitously in all tissues (McGrail et al., 1991). By mutating zebrafish orthologs of Atp1a3, we therefore could test the neuronal-specific role of the Na+,K+-ATPase in sleep.

Mutations in both zebrafish Atp1a3 orthologs increased waking locomotor behaviour during the day. However, only mutations in atp1a3a, which is expressed brain-wide, but not in atp1a3b, which is expressed in more restricted brain regions, led to changes in night-time sleep. The atp1a3a mutants have a larger sleep reduction than dmistvir, dmisti8, or fxyd1Δ28 mutants, which is expected since loss of a pump subunit should have a larger effect than the loss of a modulatory subunit, as has been shown for other ion channels (Cirelli et al., 2005; Wu et al., 2014). Autosomal dominant missense mutations leading to loss of function in ATP1A3 cause movement disorders such as rapid-onset dystonia parkinsonism and childhood alternating hemiplegia (recurrent paralysis on one side) in humans (Canfield et al., 2002; Heinzen et al., 2014), while loss of function mutations in Atp1a3 result in generalised seizures and locomotor abnormalities, including hyperactivity, in mice, which was not observed in zebrafish (Clapcote et al., 2009; Hunanyan et al., 2015; Ikeda et al., 2013; Kirshenbaum et al., 2011; Sugimoto et al., 2014). A very high prevalence of insomnia was recently reported in patients with childhood alternating hemiplegia, some of which harboured mutations in Atp1a3 (Kansagra et al., 2019), consistent with our observations that insomnia at night is a direc behavioural consequence of atp1a3a mutation in zebrafish. Since zebrafish atp1a3a mutants phenocopy the insomnia and hyperactivity phenotypes observed in patients, small molecule screens aimed at ameliorating zebrafish atp1a3a mutant phenotypes may be a promising approach for the rapid identification of new therapies for the management of this disease (Hoffman et al., 2016; Rihel et al., 2010).

Together, the night-specific sleep phenotypes of dmist, fxyd1, and atp1a3a mutants point to a role for the Na+,K+-ATPase in boosting sleep at night. How might the alpha3 catalytic subunit of the Na+/K+ pump regulate sleep, and how could Dmist be involved? We found that Dmist is required for proper maintenance of brain intracellular Na+ levels at night but not during the day, mirroring the timing of sleep disruption in dmisti8/i8 animals. This suggests that the decreased night-time sleep of dmist mutants is due to a specific requirement for Dmist modulation of the Na+/K+ pump at night. However, we cannot exclude the possibility that Dmist’s function is required in only a subset of critical sleep/wake regulatory neurons during the day that then influence behaviour at night, such as the wake-active, sleep-homeostatic regulating serotonergic neurons of the raphe (Oikonomou et al., 2019) or wake-promoting Hcrt/orexin neurons (Li et al., 2022). Another possibility is that disruption of proper establishment of the Na+ electrochemical gradient in dmist mutant neurons leads to dysfunction of various neurotransmitter reuptake transporters, including those for glycine, GABA, glutamate, serotonin, dopamine, and norepinephrine, which rely on energy from the Na+ gradient to function (Kristensen et al., 2011).

A third possibility is that Dmist and the Na+,K+-ATPase regulate sleep not by modulation of neuronal activity per se but rather via modulation of extracellular ion concentrations. Recent work has demonstrated that interstitial ions fluctuate across the sleep/wake cycle in mice. For example, extracellular K+ is high during wakefulness, and cerebrospinal fluid containing the ion concentrations found during wakefulness directly applied to the brain can locally shift neuronal activity into wake-like states (Ding et al., 2016). Given that the Na+,K+-ATPase actively exchanges Na+ ions for K+, the high intracellular Na+ levels we observe in atp1a3a and dmist mutants is likely accompanied by high extracellular K+. Although we can only speculate at this time, a model in which extracellular ions that accumulate during wakefulness and then directly signal onto sleep-regulatory neurons could provide a direct link between Na+,K+ ATPase activity, neuronal firing, and sleep homeostasis. Such a model could also explain why disruption of Fxyd1 in non-neuronal cells also leads to a reduction in night-time sleep.

In addition to decreased night-time sleep, we also observed that dmist mutants have an exaggerated sleep rebound response following the high, widespread neuronal activity induced by the GABA-receptor antagonist, PTZ. Since both Atp1a3a and Dmist were essential for re-establishing proper brain intracellular Na+ levels following PTZ exposure (Figure 6A), we speculate that the exaggerated sleep rebound is a consequence of increased neuronal depolarization due to defective Na+ pump activity. This is consistent with our previous observations that the intensity of brain-wide neuronal activity impacts the magnitude of subsequent sleep rebound via engagement of the Galanin sleep-homeostatic output arm (Reichert et al., 2019). Why does loss of dmist lead to both decreased night-time sleep and increased sleep rebound in response to exaggerated neuronal activity during the day? One possibility is that Na+/K+ pump complexes made up of different alpha and beta subunits may be differentially required for maintaining Na+ homeostasis under physiological conditions and have different affinities for (or regulation by) Dmist. For example, the Atp1a1 subunit is considered the Na+/K+ pump workhorse in neurons, while Atp1a3, which has a lower affinity for Na+ ions, plays an essential role in repolarizing neurons when Na+ rapidly increases during high levels of neuronal activity, such as after a seizure (Azarias et al., 2013). If Dmist preferentially interacts with Atp1a3a subunit, with which the non-additive effect of dmist and atp1a3a mutation on sleep is consistent, day-time sleep-related phenotypes in dmist mutants might be uncovered only during physiological challenge. Conversely, neurons may be more dependent on Atp1a3a and Dmist for sodium homeostasis at night due to changes in Na+/K+ pump composition, Dmist interactions, or ion binding affinities. For example, activity of the Na+/K+ pump can be modulated by the circadian clock (Damulewicz et al., 2013; Nakashima et al., 2018), changes in substrate availability, including ATP (reviewed in Therien and Blostein, 2000), or hormones (Ewart and Klip, 1995). Teasing out how Dmist modulation of the Na+/K+ pump changes across the day-night cycle, and in which neurons Dmist’s function may be particularly important at night, will require future investigation.

In conclusion, through a genetic screening strategy in zebrafish, we have identified a novel brain expressed gene that encodes a small transmembrane protein regulator of night-time sleep and wake behaviours. Future work will be required to uncover the precise signalling dynamics by which Dmist regulates the Na+,K+-ATPase and sleep.

Supporting information

eLife Assessment

This study offers new fundamental information on a role for the sodium/potassium pump in sleep regulation. Elegant methods were used to provide compelling evidence supporting the claim. The work will be of interest to sleep researchers in zebrafish as well as in other species.

Joint Public Review:

Barlow et al performed a viral insertion screen in larval zebrafish for sleep mutants. They identify a mutant named dreammist (dmist) that displayed defects in sleep, namely, decreased sleep both day and night, accompanied by increased activity. They find that dmist encodes a previously uncharacterized single-pass transmembrane protein that shows structural similarity to Fxyd1, a Na+K+-ATPase regulator. They go on to show that genetic manipulations of either FXYD1 or the Na/K pump also reduce sleep. They use pharmacology and sleep deprivation experiments to provide further evidence that the NA/K pump regulates intracellular sodium and rebound sleep.

This study provides additional evidence for the important role of membrane excitability in sleep regulation. The conclusions of this paper are mostly well supported by data, with the following strengths and weaknesses as described below.

Strengths:

Elegant use of CRISPR knockout methods to disrupt multiple genes that help establish the importance of regulating Na+K+-ATPase function in sleep.

Data are mostly clearly presented.

Double mutant analysis of dmist and atp1a3a help establish an epistatic relationship between these proteins.

Weaknesses:

The authors emphasize the role of increased cellular sodium. It will be interesting to also see the consequences of perturbating potassium. The potassium channel shaker has been previously identified as a critical sleep regulator in Drosophila.

Author Response

The following is the authors’ response to the original reviews.

We thank the reviewers for their comments. We have now addressed all the comments in a revised version of the manuscript, which we believe has strengthened our paper.

1. Introduction LINE 60: the authors cite Funato et al 2016 as the paper first describing a role for SIk3 in sleep regulation. In fact, the role for this kinase was first identified nearly a decade earlier in C. elegans (Van der Linden et al, Genetics 2008 PMID 18832350).

Thank you for pointing us to this reference. Van der Linden et al. demonstrated that the C. elegans homolog of Sik3 (KIN-29) regulates satiety quiescence, in which worms stop moving following feeding on high quality food. However, as pointed out in Trojanowski and Raizen “Call it Worm Sleep” (2016), not all of the behavioral criteria for sleep has been applied to C. elegans satiety quiescence, and we cannot find any references that unequivocally demonstrate satiety quiescence is a sleep state. As McClanahan et al., (2020) show, quiescent states following mild sensory arousal do not fulfill the sleep criteria of changes in arousal threshold and homeostatic regulation, so not all quiescent states in C. elegans are sleep. Then again Grubbs et al, 2020 does demonstrate that KIN29 regulates both developmentally timed and stress induced sleep states in worms, suggesting that the observations in Van der Linden were ahead of its time and these behavioral states are possibly inter-related. We believe, though, that our line “the roles of… SIK3 kinase in modulating sleep homeostasis in mice (Funato et al. 2016) were identified in genetic screens” remains accurate.

1. Introduction LINE 71: remove the word "known" from "...while some known human sleep/wake regulators, such as the...")

Good idea. Done.

1. I was confused regarding Supplemental data 1 describing the genes they targeted with their forward genetic screen. Am I understanding correctly from the "Summary stats" tab that 702 fish lines with virus insertions were screened behaviorally? In Figure S1, it looks like about 60 are shown in the histograms but in the text (in the Discussion) they say 25 were screened. Were all the genes listed under the Excel tabs (GPCRs, channels, etc) tested? Or was just a subset tested? Where are the sleep data for these lines? Negative results may be relevant to their manuscript since they listed (tested??) a number of ion channel genes under tab "channels" which appear to NOT have a sleep phenotype.

We apologize for the confusion on these points. As highlighted in the legend to Supplementary Figure S1, we had planned a screening strategy with the following pipeline: Candidate mammalian gene → Zebrafish ortholog → ID viral insertion from “Zenemark” library → grow viral insertion lines from frozen sperm→ phenotype F3 heterozygous and homozygous mutant generation. Unfortunately, the company, Znomics, which held the Zenemark library, could not reliably reconstitute the correct live fish from the sperm library, and of the 702 lines we planned to screen, we could only screen 26 (25 was a typo) lines. We treated heterozygous and homozygous animals for each line independently, for a total of 52 screened lines in the histograms.

To make this clearer, we have edited the main text as follows (lines 104-105): “For screening, we identified zebrafish sperm samples from the Zenemark collection (Varshney et al., 2013) that harboured viral insertions in genes of interest and used these samples for in vitro fertilization and the establishment of F2 families, which we were able to obtain for 26 lines.” And lines 111-112: “While most screened heterozygous and homozygous lines had minimal effects on sleep-wake behavioural parameters (Figure S1B-S1C),”

We believe it is important to include the full set of Supplementary Data 1, even though the vast majority of these candidate lines were not tested.

1. Results LINE 117: remove the word "prominent", which is subjective, from the sentence "...showed a prominent decrease in sleep during the..."

Good point. Done.

1. LINES 185-186: did you see any circadian variation in your dmist:GFP protein abundance or localization? Protein trafficking has been described as a mechanism of circadian regulation of excitability.

For practical reasons, we imaged the membrane localization of Dmist:GFP in plasmidinjected embryos at 90% epiboly, which is about 9 hours after fertilization and when the cells remain large and in a relatively flat epithelium. Thus, we could not follow circadian fluctuations in abundance or localization. For circadian studies, we believe the best method will be to raise an antibody that recognizes Dmist.

1. LINE 203: does the GFP-tagged Dmist rescue the loss-of-function phenotype? This is relevant to Figure 2E. it is also relevant to the issue of structure-function. If it rescues, then the C-terminus may not be essential to protein function.

As noted, for practical reasons, we observed Dmist-GFP only transiently at early stages of development, expressed using a strong, ubiquitous promoter. A rescue experiment is a good idea for future experiments, where we carefully control the expression of Dmist in neurons.

1. LINE 220: explain what you mean by "...consistent with nonsense-mediated decay." and/or give a reference.

In zebrafish and other species including humans, mutant transcripts that have premature stop codons often undergo “nonsense mediated decay”, whereby the expression levels are largely reduced (Wittkopp et al., 2009). In the zebrafish community, this is often used as secondary evidence of a loss of function mutation, as relatively few antibodies are available to directly observe zebrafish proteins. We have added a reference that describes this phenomenon (Wittkopp et al., 2009).

1. LINE 225: define "LME model"

Now reads: “Linear mixed effects (LME).”

1. LINES 227-229: could the vir/vir phenotype be explained by specific effects on protein structure? could vir/vir be a gain-of-function allele?

We can’t rule this out formally, and vir/+ animals do show some sleep phenotypes, albeit weaker than those of vir/vir animals (Figure 1G). However, it is not uncommon for heterozygous mutants to show significant phenotypes that are weaker than those of their homozygous mutant siblings, and the strong suppression of dmist expression by the viral insertion (which is located in the dmist intron) is more consistent with a hypomorphic loss-of-function phenotype for the vir allele.

1. LINES 229-230: I don't quite follow the argument for pursuing further studies only of i8/i8. i8/i8 seems to also be a hypomorphic allele based on your qPCR data.

First, the dmist viral line was generated by an insertional mutagenesis method followed by sequencing, and each line has multiple other inserts in a background that does not match the background of the other animals reported in this paper. Second, the dmist vir allele is an insertion in the intron, leading to reduced, but not complete loss of expression. In contrast, the i8 allele was generated on the same background strain as our other existing and newly reported lines. Moreover, our i8 line is likely a loss-of-function allele and not a hypomorph. Yes, dmist expression is reduced in the i8 allele; however, this is likely due to nonsense mediated decay of dmist mRNA. The mutation introduces a frameshift in the dmist coding sequence, and as a result the amino acid sequence of the protein is altered after the N-terminal signal sequence.

1. LINES 241-243: grammar.

Fixed

1. LINE 245: define "JackHMMR iterative search"

We’ve added the phrase: “and seeding a hidden Markov model iterative search (JackHMMR)”

1. LINE 246 is missing the word "we" prior to "...found distant homology between..."

Added

1. LINE 301: show data demonstrating deviation from Mendelian ratios. Also, comment on meaning of such data (embryonic lethality??).

We have added this data in the line (301):

“atp1a3b mutant larvae were not obtained at Mendelian ratios (55 wild type [52.5 expected], 142 [105] atp1a3b+/-, 13 [52.5] atp1a3b-/-; p<0.0001, Chi-squared) suggesting some impact on early stages of development leading to lethality.”

1. Discussion LINES 362-372: This paragraph seems to be of only tangential relevance to the paper. Consider removing.

Our screening strategy was a large-scale reverse genetic screen, but the number of lines was limited by the technical issues described above. We think it is important to mention that the strategy, if employed today, could benefit from newer technologies.

1. Discussion. Another model is that Dmist and NaK pump have a developmental effect. Arguing against this developmental model is the Oubain expt.

This is an important point. We’ve added the line (454:457): “We also cannot exclude a role for Dmist and the Na+/K+ pump in developmental events that impact sleep, although our observation that ouabain treatment, which inhibits the pump acutely after early development is complete, also impacts sleep, argues against a developmental role.”

1. FIGURE 1G: Are these significance cut offs corrected for multiple comparisons?

Yes, all the data is corrected for multiple comparisons.

1. performing neuronal activity measures, either via neural activity imaging or phospho-ERK labeling in different mutants at day or night conditions, to determine whether baseline neuronal activity brain-wide or in specific brain regions are altered.

These are excellent experiments that we plan to perform in the future.

1. Please check all Figure numbers for accuracy.

We have double checked these.

1. The authors emphasize the role of increased cellular sodium, but equally plausibly, the phenotypes could be due to decreased cellular potassium. The potassium channel shaker has been previously identified as a critical sleep regulator in Drosophila.

We completely agree. We would like to highlight that we did devote an entire paragraph to the possibility of changes in extracellular potassium in the discussion: “A third possibility is that Dmist and the Na+,K+-ATPase regulate sleep not by modulation of neuronal activity per se but rather via modulation of extracellular ion concentrations. Recent work has demonstrated that interstitial ions fluctuate across the sleep/wake cycle in mice. For example, extracellular K+ is high during wakefulness, and cerebrospinal fluid containing the ion concentrations found during wakefulness directly applied to the brain can locally shift neuronal activity into wake-like states (Ding et al., 2016). Given that the Na+,K+-ATPase actively exchanges Na+ ions for K+ , the high intracellular Na+ levels we observe in atp1a3a and dmist mutants is likely accompanied by high extracellular K+. Although we can only speculate at this time, a model in which extracellular ions that accumulate during wakefulness and then directly signal onto sleep-regulatory neurons could provide a direct link between Na+,K+ ATPase activity, neuronal firing, and sleep homeostasis. Such a model could also explain why disruption of fxyd1 in non-neuronal cells also leads to a reduction in night-time sleep.”

We also agree that Shaker may be an important component of this sleep regulatory mechanism. Indeed, we previously showed that another potassium channel in zebrafish regulates sleep (Rihel et al., 2010).

We have emphasized sodium homeostasis in our title and paper only because we were able to directly observe intracellular sodium levels, so we are confident that these have been altered in our mutants. We can only presume that potassium levels have also been altered, but we could not directly observe this.

1. The similar phenotype between dmist and Fxyd1 in sleep reduction yet very different expression patterns, with dmist being mostly neuronal while fxyd1 being mostly non-neuronal, raise many possible questions: 1) are the sleep phenotypes due to neuronal Na/K imbalance? Or 2) Are the sleep phenotypes due to extracellular Na/K imbalance? Or 3) both? Some feasible experiments may help achieve a better mechanistic understanding of the observed sleep defects.

Yes, we think these are excellent studies for future work. As noted in the previous point (20), we did discuss the possibility that changes to extracellular potassium might be a parsimonious explanation for the similar phenotypes of fxyd1 and dmist mutants.

Future experiment suggestions (not required)

1. Perform a double mutant analysis of fxyd1 and atp1a3a, to determine whether an epistatic relationship similar to that of dmist and atp1a3a is observed in the case of fxyd1 and atp1a3a.

This is a great experiment that we will do in the future. Unfortunately, the fxyd1 mutant had been sperm frozen during the COVID-19 pandemic, so we cannot do this experiment at this time.

1. Given the differences in the sleep phenotypes between vir/vir and i8/i8 mutants, would be informative to see the phenotype of the vir/i8 trans-heterozygote.

This is also a good experiment to perform in the future. Since obtaining the cleaner i8 allele, the dmistvir/vir lines were sperm frozen.

Version 2

Abstract

Abstract

Sleep is a nearly universal feature of animal behaviour, yet many of the molecular, genetic, and neuronal substrates that orchestrate sleep/wake transitions lie undiscovered. Employing a viral insertion sleep screen in larval zebrafish, we identified a novel gene, dreammist (dmist), whose loss results in behavioural hyperactivity and reduced sleep at night. The neuronally expressed dmist gene is conserved across vertebrates and encodes a small single-pass transmembrane protein that is structurally similar to the Na+,K+-ATPase regulator, FXYD1/Phospholemman. Disruption of either fxyd1 or atp1a3a, a Na+,K+-ATPase alpha-3 subunit associated with several heritable movement disorders in humans, led to decreased night-time sleep. Since atpa1a3a and dmist mutants have elevated intracellular Na+ levels and non-additive effects on sleep amount at night, we propose that Dmist-dependent enhancement of Na+ pump function modulates neuronal excitability to maintain normal sleep behaviour.

Introduction

The ability of animals to switch between behaviourally alert and quiescent states is conserved across the animal kingdom (Cirelli, 2009; Joiner, 2016). Fundamental processes that govern the regulation of sleep-like states are shared across species, such as the roles of circadian and homeostatic cues in regulating the time and amount of sleep, stereotyped postures, heightened arousal thresholds, and the rapid reversibility to a more alert state (Joiner, 2016). The near ubiquity of sleep implies that it serves ancient functions and is subject to conserved regulatory processes. However, many key molecular components that modulate sleep and wake states remain undiscovered.

Over the past two decades, investigations into sleep and arousal states of genetically tractable model organisms, such as Drosophila melanogaster, C. elegans, and Danio rerio (zebrafish) have uncovered novel molecular and neuronal components of sleep regulation through gain- and loss-of-function genetic screens (reviewed in Barlow and Rihel, 2017; Sehgal and Mignot, 2011). The power of screening approaches is perhaps best exemplified by the first forward genetic sleep screen, which identified the potassium channel shaker as a critical sleep regulator in Drosophila (Cirelli et al., 2005). This result continues to have a lasting impact on the field, as not only did subsequent sleep screening efforts uncover the novel Shaker regulator sleepless, (Koh et al., 2009), but investigations into Shaker’s beta subunit Hyperkinetic ultimately revealed a critical role for this redox sensor linking metabolic function to sleep (Bushey et al., 2007; Kempf et al., 2019).

Disparate screening strategies across model organisms continue to unveil novel sleep modulators in both invertebrate and vertebrate model systems. For example, the roles of RFamide receptor DMSR-1 in stress-induced sleep in C. elegans (Iannacone et al., 2017) and SIK3 kinase in modulating sleep homeostasis in mice (Funato et al., 2016) were identified in genetic screens. Moreover, a gain of function screening strategy in Drosophila revealed the novel sleep and immune regulator, nemuri (Toda et al., 2019), and a zebrafish overexpression screen uncovered the secreted neuropeptides neuromedin U and neuropeptide Y, which decrease and increase sleep, respectively (Chiu et al., 2016; Singh et al., 2017). The success of screening strategies in revealing novel sleep-wake regulatory genes suggests that more sleep signals likely remain to be discovered.

One of the lessons from these genetic screens is that many of the uncovered genes play conserved roles across species. For example, Shaker also regulates mammalian sleep (Douglas et al., 2007) and RFamides induce sleep in worms, flies, and vertebrates (Lee et al., 2017; Lenz et al., 2015). Nevertheless, not every invertebrate sleep-regulatory gene has a clear vertebrate homolog, while some human sleep/wake regulators, such as the narcolepsy- associated neuropeptide hypocretin/orexin (Chemelli et al., 1999; Lin et al., 1999; Peyron et al., 2000; Sakurai, 2013), lack invertebrate orthologs. Therefore, genetic sleep screens in vertebrates are likely to provide added value in uncovering additional regulatory components required to control the initiation and amount of sleep in humans.

While sleep screening in mammals is feasible (Funato et al., 2016), it remains an expensive and technically challenging endeavour. With its genetic tractability, availability of high- throughput sleep assays (Rihel and Schier, 2013), and conserved sleep genetics, such as the hypocretin, melatonin, locus coeruleus, and raphe systems (Gandhi et al., 2015; Singh et al., 2015; Oikonomou et al., 2019; Prober et al., 2006), the larval zebrafish is an attractive vertebrate system for sleep screens. We took advantage of a collection of zebrafish lines that harbour viral-insertions in >3500 genes (Varshney et al., 2013) to perform a targeted genetic screen. We identified a short-sleeping mutant, dreammist, with a disrupted novel, highly conserved vertebrate gene that encodes a small single pass transmembrane protein. Sequence and structural homology to the Na+/K+ pump regulator FXYD1/Phospholemman suggests that Dreammist is a neuronal-expressed member of a class of sodium pump modulators that is important for regulating sleep-wake behaviour.

Results

Reverse genetic screen identifies dreammist, a mutant with decreased sleep

We used the ‘Zenemark’ viral-insertion based zebrafish gene knock-out resource (Varshney et al., 2013) to perform a reverse genetic screen to identify novel vertebrate sleep genes. This screening strategy offers several advantages compared to traditional chemical mutagenesis- based forward genetic screening approaches. First, unlike chemical mutagenesis, which introduces mutations randomly, viral insertions tend to target the 5’ end of genes, typically causing genetic loss of function (Sivasubbu et al., 2007). Second, because the virus sequence is known, it is straightforward to map and identify the causative gene in mutant animals. Finally, since viral insertions in the Zenemark collection are already mapped and sequenced, animals harbouring insertions within specific gene classes can be selected for testing (Figure S1A). This allowed us to prioritise screening of genes encoding protein classes that are often linked to behaviour, such as G-protein coupled receptors, neuropeptide ligands, ion channels, and transporters (Supplemental Data 1).

For screening, we identified zebrafish sperm samples from the Zenemark collection (Varshney et al., 2013) that harboured viral insertions in genes of interest and used these samples for in vitro fertilization and the establishment of F2 families, which we were able to obtain for 26 lines. For each viral insertion line, clutches from heterozygous F2 in-crosses were raised to 5 days post-fertilisation (dpf) and tracked using videography (Figure S1A) to quantify the number and duration of sleep bouts (defined in zebrafish larvae as inactivity lasting 1 minute or longer; Prober et al., 2006) and waking activity (time spent moving per active bout) over 48 hours. The genotypes of individual larvae were determined by PCR after behavioural tracking, with each larva assigned as wild type, heterozygous, or homozygous for a given viral insertion to assess the effect of genotype on sleep/wake behaviour. While most screened heterozygous and homozygous lines had minimal effects on sleep-wake behavioural parameters (Figure S1B-S1C), one homozygous viral insertion line, 10543/10543, had a reduction in daytime sleep (Figure S1B) and an increase in daytime waking activity (Figure S1C) relative to their wild type sibling controls. We re-named this 10543 viral insertion line dreammist (dmist).

In follow-up studies, we observed that animals homozygous for the viral insertion at this locus (dmistvir/vir) showed a decrease in sleep during the day and a trend to sleep less at night compared to their wild-type siblings (dmist+/+) (Figure 1A). dmist mutants had an almost 50% reduction in the average amount of daytime sleep (Figure 1C) due to a decrease in the number of sleep bouts (Figure 1D), whereas the sleep bout length at night was significantly reduced (Figure 1E). dmistvir/vir larvae also exhibited significantly increased daytime waking activity, which is the locomotor activity while awake (Figure 1B, 1F). Because Zenemark lines can contain more than one viral insertion (17.6% of lines have ≥2 insertions; Varshney et al 2013), we outcrossed dmistvir/+ fish to wild-type fish of the AB-TL background and re-tested dmist mutant fish over several generations. Normalising all the behavioural parameters to dmist+/+ controls with a linear mixed effects (LME) model showed consistent sleep changes in dmistvir/vir fish over 5 independent experiments (Figure 1G). The dmistvir/vir larvae consistently show a more than 50% decrease in sleep during the day due to a significant reduction in the number and duration of sleep bouts, as well as a large increase in waking activity (Figure 1G). The dmistvir/vir mutants also had a significant reduction in sleep at night compared to wild type siblings (Figure 1G). These effects on sleep and wakefulness are not due to alterations in circadian rhythms, as behavioural period length in fish that were entrained and then shifted to free-running constant dark conditions was unaffected in dmistvir/vir compared to wild-type sibling larvae (Figure S2A-S2C).

A viral insertion mini-screen identifies a short-sleeping mutant, dreammist.

A-B) Mean± SEM sleep (A) and waking activity (B) of progeny from dmistvir/+ in-cross from original screen. White blocks show day (lights on) and grey blocks show night (lights off). Data is combined from 2 independent experiments. n indicates the number of animals.

C-F) Analysis of sleep/wake architecture for the data shown in (A, B). C) Quantification of total sleep across two days and nights shows decreased day and night sleep in dmistvir/vir. Analysis of sleep architecture reveals fewer sleep bouts during the day (D) and shorter sleep bouts at night (E) in dmistvir/vir compared with sibling controls. F) Daytime waking activity is also increased in dmistvir/vir. The black lines show the mean ± SEM, except in E, which labels the median ± SEM. *p<0.05, **p<0.01, ***p<0.001; ns p>0.05; one-way ANOVA, Tukey’s post hoc test.

G) Combining 5 independent experiments using a linear mixed effects model with genotype as a fixed effect and experiment as a random effect reveals dmistvir/vir larvae have decreased total sleep and changes to sleep architecture during both the day and night compared to dmist+/+ siblings. Plotted are the genotype effect sizes (95% confidence interval) for each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05,

**p<0.01, ***p<0.001, ns p>0.05. n indicates the number of animals.

The dmist gene encodes a novel, small transmembrane protein

Having identified a sleep mutant, we next sought to investigate the target gene disrupted by the viral insertion. Line 10543 (dmistvir) was initially selected for screening due to a predicted disruption of a gene encoding a serotonin transporter (slc6a4b) on chromosome 5. However, mapping of the dmist viral insertion site by inverse-PCR and sequencing revealed that the virus was instead inserted into the intron of a small two-exon gene annotated in the Zv6 genome assembly as a long intergenic non-coding RNA (lincRNA; gene transcript ENSDART00000148146, gene name si:dkey234h16.7), which lies approximately 6 kilobases (kb) downstream of the slc6a4b gene in zebrafish. At least part of this region is syntenic across vertebrates, with a small two-exon gene identified adjacent to the genes ankrd13a and GIT in several vertebrates, including human and mouse (Figure 2A). Amplifying both 5’ and 3’ ends of zebrafish si:dkey234h16.7 and mouse E13.5 1500011B03-001 transcripts with Rapid Amplification of cDNA ends (RACE) confirmed the annotated zebrafish and mouse transcripts and identified two variants with 3’ untranslated regions (3’UTR) of different lengths in zebrafish (Figure S3B). To test whether the viral insertion in dmistvir/vir disrupts expression of si:dkey234h16.7 or neighbouring genes, we performed quantitative analysis of gene transcript levels in wild type and mutant dmist larvae by RT-qPCR. This revealed that the dmist viral insertion caused a more than 70% reduction in the expression of si:dkey234h16.7 while the expression of the most proximal 5’ or 3’ flanking genes, slc6a4b_Dr and ankrd13a_Dr, were unaffected (Figure 2B and S3A). Since this reduced expression is most consistent with si:dkey234h16.7 being the causal lesion of the dmist mutant sleep phenotype, we renamed this gene dreammist (dmist).

dmist encodes a conserved vertebrate single pass transmembrane protein.

A) dmist mutants harbour a viral insertion in the 1st intron of si:key-234h16.7. dmist is syntenic with Ankrd13 and GIT orthologs in mouse, human, and zebrafish.

B) RT-qPCR of dmist (red) show reduced expression of dmist and not the 5’ and 3’ flanking zebrafish genes, slc6a4b (cyan) and ankrd13a (blue), in dmistvir/vir larvae compared to dmistvir/+ and dmist+/+ siblings. **p<0.01, *p<0.05; ns p>0.05; one-way ANOVA, Tukey’s post-hoc test. Data shows mean ± SEM normalized to the wild type mean.

C) dmist_Dr contains an open reading frame encoding a 70 amino acid protein that is conserved across vertebrates. All identified homologues have a predicted signal peptide sequence (magenta line), signal peptide cleavage site (magenta circle), and predicted transmembrane domain (grey), with additional highly conserved C-terminal motifs (blue lines). Identical amino acids in all species are shown in black; similar amino acids (80-99% conserved across species) are shown in grey.

D) In situ hybridisation using a dmist antisense probe reveals dmist is maternally deposited as it is detected at the 2-cell stage. At 24 hpf expression is restricted to regions containing neuronal precursors, and at 5 dpf expression is widespread throughout the brain. Tel, telencephalon; Dien, diencephalon; R1-6, rhombomeres 1-6; A, anterior; P, posterior. Scale bars= 0.5 mm (2 cell and 24 hpf), 0.1 mm (5 dpf).

E-F) Representative confocal image of 90% epiboly embryo co-injected at the 1-cell stage with mRNA encoding membrane-RFP (magenta) and a plasmid encoding either C-terminal tagged Dmist-GFP (E, green) or DmistA22W-GFP (F, green). Scale bar= 25 μm.

Computational predictions indicated that the dmist transcripts contain a small open reading frame (ORF) encoding a protein of 70 amino acids (aa) (Figure 2C). Querying the human and vertebrate protein databases by BLASTp using the C-terminal protein sequence of Dmist identified orthologs in most vertebrate clades, including other species of teleost fish, birds, amphibians, and mammals (Figure 2A, C). All identified orthologs encoded predicted proteins with an N-terminal signal peptide sequence and a C-terminal transmembrane domain (Figure 2C). The peptide sequence identity across orthologs ranged from 38 to 84%, with three peptide motifs (QNLV, CVYKP, RRR) showing high conservation across all vertebrates, and high similarity for many additional residues (Figure 2C, Figure S3D). Additional searches by tBLASTn failed to identify any non-vertebrate dmist orthologs. In summary, we found that the dreammist gene, the expression of which is disrupted in dmistvir/vir fish with sleep phenotypes, encodes a protein of uncharacterized function that is highly conserved across vertebrates at both the genomic and molecular levels.

Genetic molecular analysis of dmist expression in zebrafish and mouse

Because the viral insertion disrupts dmist throughout the animal’s lifetime, we examined both the developmental and spatial expression of dmist to assess when and where its function may be required for normal sleep. Using the full-length transcript as a probe (Figure S3B), we performed in situ hybridization across embryonic and larval zebrafish development. Maternally deposited dmist was detected in early embryos (2-cell stage) prior to the maternal to zygotic transition (Giraldez et al., 2006) (Figure 2D). Consistent with maternal deposition of dmist transcripts, inspection of the 3’ end of the dmist gene revealed a cytoplasmic polyadenylation element (‘TTTTTTAT’; Supplemental Information 2) that is required for zygotic translation of maternal transcripts (Villalba et al., 2011). At 24 hpf, transcripts were detected in regions that form the embryonic brain, such as ventral telencephalon, diencephalon and cerebellum, and in the developing eye (Figure 2D, S3C). By 5 dpf, dmist transcripts were detected throughout the brain (Figure 2D). To test whether dmist transcripts are under circadian regulation, we performed RT-qPCR in fish that were entrained and then shifted to free-running constant dark conditions. In contrast with the robust 24-hr rhythmic transcription of the circadian clock gene per1, we did not detect any changes in dmist expression throughout the 24 hour circadian cycle (Figure S2D).

Consistent with brain expression in larval zebrafish, we identified the expression of Dmist_Mm in a published RNAseq dataset of six isolated cell types from mouse cortex (Zhang et al., 2014). We confirmed that Dmist_Mm is specifically enriched in neurons by hierarchical clustering of all 16,991 expressed transcripts across all six cells types, which demonstrated that Dmist_Mm co-clusters with neuronal genes (Figure S3E). Pearson correlation of Dmist_Mm with canonical markers for the six cell types showed that Dmist_Mm expression is highly correlated with other neuronal genes but not genes associated with microglia, oligodendrocytes, or endothelia. This result indicates that dmist is specifically expressed in neurons in both zebrafish and mouse (Figure S3F).

Dmist localises to the plasma membrane

Although the dmist gene encodes a conserved ORF with a predicted signal peptide sequence and transmembrane domain (Figure 2C; Figure S3G-I), we wanted to confirm this small peptide can localise to the membrane and if so, on which cellular compartments. To test these computational predictions, we transiently co-expressed GFP-tagged Dmist (C-terminal fusion) with a marker for the plasma membrane (myr-Cherry) in zebrafish embryos. Imaging at 90% epiboly revealed Dmist-GFP localised to the plasma membrane (Figure 2E). Conversely, introducing a point mutation into Dmist’s signal peptide cleavage site (DmistA22W-GFP) prevented Dmist from trafficking to the plasma membrane, with likely retention in the endoplasmic reticulum (Figure 2F). Together, these data indicate that Dmist localises to the plasma membrane despite its small size, as computationally predicted.

CRISPR/Cas9 generated dmisti8 mutant exhibits decreased night-time sleep

dmist expression was reduced by 70% in the viral insertion line, suggesting that dmistvir is a hypomorphic allele. To confirm that the sleep phenotypes observed in dmistvir/vir animals are due to the loss of Dmist function, we used CRISPR/Cas9 to create an independent dmist loss of function allele. We generated a zebrafish line in which the dmist gene contains an 8 bp insertion that causes a frameshift and early stop codon (dmisti8, Figure 3A). The dmisti8 allele is predicted to encode a truncated protein lacking the complete signal peptide sequence and transmembrane domain (Figure 3B), indicating this is likely a null allele. RT-qPCR showed that dmist transcript levels were 60% lower in dmisti8/i8 fish compared to wild type siblings, consistent with nonsense-mediated decay (Figure S4A, B) (Wittkopp et al., 2009).

CRISPR-generated dmist mutants sleep less and are hyperactive at night.

A) CRISPR/Cas9 targeting of the first exon of dmist resulted in an 8 bp insertion (dmisti8) (grey line) within the coding sequence, leading to an early stop codon (red line with *). Guide RNA target sequence and PAM sequence are shown as black bars. The sequence that is deleted in the mutant is indicated with a red bar.

B) Predicted Dmisti8 peptide sequence lacks most of the N-terminal signal peptide sequence (magenta) and the full C-terminus.

C-D) Representative 48 hr traces of mean ± SEM sleep (C) and waking activity (D) shows decreased sleep and increased waking activity at night for dmisti8/i8 fish compared to dmisti8/+ and dmist+/+ siblings. n=number of fish.

E-H) Analysis of sleep/wake architecture of the experiment depicted in (C, D) indicates that dmisti8/i8 larvae sleep less at night (E) due to fewer sleep bouts (F). Sleep bout length is unchanged (G). Waking activity is also increased in dmisti8/i8 fish (H). The black line represents the mean ± SEM except for G, which is the median ± SEM. *p<0.05, **p<0.01, ***p<0.001; One-way ANOVA, Tukey’s post hoc test.

I) Combining 5 independent experiments with a linear mixed effects model reveals dmisti8/i8 fish sleep less at night due to fewer sleep bouts and also show increased waking activity at night. Plotted are the genotype effect sizes (95% confidence interval) for each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05,

**p<0.01, ***p<0.001, ns p>0.05.

J-K) (J) Adult dmisti8/i8 fish have a higher mean swim speed compared to their wild type siblings at night. Data in (J) is quantified at night in (K). (J, K) show mean ± SEM. *p<0.05, one-way ANOVA.

L) Cumulative probability distribution of all night-time swim bout speeds in adult fish. The dashed lines show the half max (0.5 probability) for each curve. *p<0.05 for dmisti8/i8 fish compared to wild type siblings; Kolmogorov-Smirnov test.

We next assessed the sleep and activity patterns of dmisti8/i8 fish. As seen in exemplar individual tracking experiments, dmisti8/i8 larvae sleep less at night due to fewer sleep bouts and also show an increase in waking activity relative to wild type and heterozygous mutant siblings (Figure 3C-H). This significant night-time reduction in sleep and increase in hyperactivity is also apparent when combining 5 independent experiments with a linear mixed effects (LME) model to normalize behaviour across datasets (Figure 3I). Although dmistvir/vir larvae also sleep less at night (Figure 1G), the large day-time reduction in sleep observed in dmistvir/vir larvae is absent in dmisti8/i8 animals, perhaps due to differences in genetic background that affect behaviour. Because the dmistvir is likely a hypomorphic allele, we focused subsequent experiments on the CRISPR-generated dmisti8/i8 larvae.

To test whether the increased night-time activity of dmisti8/i8 mutants persists in older animals, we raised dmisti8/i8 mutants with their heterozygous and wild type siblings to adulthood in the same tank and tracked individual behaviour for several days on a 14:10 light:dark cycle. As in larval stages, dmisti8/i8 adults were hyperactive relative to both dmisti8/+ and dmist+/+ siblings, maintaining a higher mean speed at night (Figure 3J-L). This suggests that either Dmist affects a sleep/wake regulatory circuit during development that is permanently altered in dmist mutants, or that Dmist is continuously required to maintain normal levels of night-time locomotor activity.

Dmist is distantly related to the Na+/K+ pump regulator Fxyd1 (Phospholemman)

Because Dmist is a small, single pass transmembrane domain protein without any clear functional motifs and has not been functionally characterized in any species, we searched for similar peptides that might provide clues for how Dmist regulates behaviour. Using the multiple sequence alignment tool MAFFT to align the zebrafish, mouse, and human Dmist peptides (Katoh and Toh, 2010) and seeding a hidden Markov model iterative search (JackHMMR) of the Uniprot database (Johnson et al., 2010), we found distant homology between Dmist and Fxyd1/Phospholemman (Figure 4A), a small transmembrane domain peptide that regulates ion channels and pumps, including the Na+,K+-ATPase pump (Crambert et al., 2002). Dmist and Fxyd1 share 27-34% amino acid homology, including an RRR motif at the C-terminal end, although Dmist lacks a canonical FXYD sequence (Figure 4A). In addition, computational predictions using the AlphaFold protein structure database revealed structural similarities between Dmist and Fxyd1 (Jumper et al., 2021), suggesting that Dmist may belong to a class of small, single pass transmembrane ion pump regulators.

Mutation of the dmist related gene fxyd1 causes reduced sleep at night.

A) Schematic of zebrafish Dmist and Fxyd1 protein domains and alignments comparing human, mouse, and zebrafish Dmist and FXYD1 protein sequences. Black and grey shading indicate amino acid identity and similarity, respectively. The FXYD domain is indicated with a red line and the RRR motif in the C-terminus is indicated with a dark blue line.

B) CRISPR-Cas9 targeting of the 3rd exon of fxyd1 created a 28 bp deletion, resulting in a predicted truncated protein. The start codon is marked by a cyan line. Guide RNA target sequence and PAM sequence are shown as black bars. The mutant deleted sequence is indicated with a red bar.

C) In situ hybridisation of fxyd1 at 24 hpf (whole animal) and 5 dpf brain (ventral view). Anterior is to the left. Scale bar = 0.5 mm (24 hpf); 0.1 mm (5 dpf).

D-E) Representative behavioral experiment showing fxyd1Δ28 mutants have decreased night- time sleep (D) but normal waking activity at night (E).

F) Combining 5 independent experiments with a linear mixed effects model reveals fxyd1Δ28/ Δ28 larvae sleep significantly less at night due to shorter sleep bouts compared to fxyd1+/+ siblings. Plotted are the genotype effect sizes (95% confidence interval) on each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05,

**p<0.01, ***p<0.001, ns p>0.05.

Using In situ hybridisation, we found that fxyd1 is expressed in cells along the brain ventricle and choroid plexus (Figure 4C) in contrast to the neuronal expression of dmist (Figure 2D). Despite these different expression patterns, based on their sequence similarity we reasoned that Fxyd1 and Dmist may regulate the same molecular processes that are involved in sleep. To test this hypothesis, we used CRISPR/Cas9 to generate a 28 bp deletion in the third exon of the zebrafish fxyd1 gene, causing a frameshift that is predicted to encode a truncated protein that lacks the FXYD, transmembrane, and C-terminal domains (Figure 4B). Contrary to a previous report based on morpholino knockdown (Chang et al., 2012), fxyd1Δ28/Δ28 larvae were viable with no detectable defect in inflation of the brain ventricles. We therefore tested fxyd1 mutant larvae for sleep phenotypes. Like dmist mutants, fxyd1Δ28/Δ28 larvae slept less at night (Figure 4D-F). Interestingly, this sleep loss is mainly due to shorter sleep bouts (Figure 4F), indicating that fxyd1 mutants initiate sleep normally but do not properly maintain it, unlike dmist mutants, which initiate fewer night-time sleep bouts, although in both cases there is consolidation of the wake state at night (Figure 3I, 4F). Thus, despite the non-neuronal expression of fxyd1 in the brain, mutation of the gene most closely related to dmist results in a similar sleep phenotype.

The brain-wide Na+/K+ pump alpha subunit Atp1a3a regulates sleep at night

Given the similarity between Dmist and Fxyd1 and their effects on night-time sleep, we hypothesized that mutations in Na+/K+ pump subunits known to interact with Fxyd1 might also affect sleep. Consistent with this hypothesis, a low dose of the Na+/K+ pump inhibitor, ouabain, reduced night-time sleep in dose-response studies (Figure S5A). When applied in the late afternoon of 6 dpf, 1 µM ouabain decreased subsequent night-time sleep by 16.5% relative to controls, an effect size consistent with those observed in dmist mutants (Figure 5A, C). Night- time waking activity was also significantly increased after low-dose ouabain exposure (Figure 5B, D). Ouabain binds to specific sites within the first extracellular domain of Na+/K+ pump alpha subunits (Price and Lingrel, 1988), and species-specific changes to these sites confers species-specific ouabain resistance, as in the case of two naturally occurring amino acid substitutions present in the Atp1a1 subunit of mice (Dostanic et al., 2004). Alignment of the ouabain sensitive region of zebrafish and mouse Na+/K+ pump alpha subunits revealed that zebrafish Atp1a1a lacks the conserved Glutamine at position 121 (Figure 5E), suggesting that one of the other subunits with conserved ouabain-binding sites is responsible for the low dose ouabain sleep effects. We focused on the Na+/K+ pump alpha-3 subunit (Atp1a3), as this has been shown to directly interact with Fxyd1 in mammalian brain tissue (Feschenko et al., 2003). Murine Dmist expression also correlates well with the Atp1a3 distribution across 5 brain cell types in mouse (Pearson correlation coefficient = 0.63), which has the strongest correlation score with neuronal markers (Figure S5B compared to Figure S3F). In contrast, zebrafish atp1a2a is reportedly expressed in muscle at larval stages, while atp1a1b is confined to cells along the ventricle (Thisse et al., 2001).

Mutation of the Na+/K+ pump alpha subunit atp1a3a reduces sleep at night

A-B) Mean ± SEM sleep and waking activity traces of wild type larvae following exposure to 1 µM ouabain. Arrows indicate time the drug was added.

C-D) At night, sleep is significantly reduced and waking activity is significantly increased after ouabain exposure. Student’s t-test, one tailed.

E) Alignments of Na+/K+ pump alpha subunits around the ouabain binding sites. Red indicates residues that are critical for higher sensitivity to ouabain, which are present in mouse Atp1a3 but not Atp1a1.

F) In situ hybridisation of atp1a3a at 24hpf (whole animal) and 5dpf brain (ventral view). Anterior is to the left. Scale bar = 0.5 mm (24 hpf); 0.1 mm (5 dpf). A-anterior; P-posterior; D- Dorsal; V-Ventral

G) CRISPR-Cas9 targeting of the atp1a3a resulted in a 19 bp deletion that eliminates the start codon (blue) and splice junction. Guide RNA target sequence and PAM sequence are shown as black bars. Sequence that is deleted in the mutant is indicated with a red bar.

H-I) Representative behavioural experiment showing atp1a3aΔ19/Δ19 fish are hyperactive throughout the day-night cycle and have decreased sleep at night. Mean ± SEM are shown.

J) atp1a3aΔ19/ Δ19 larvae sleep less at night due to shorter sleep bouts. Plotted are the genotype effect sizes (95% confidence interval) on each parameter relative to wild type. Shading indicates day (white) and night (grey). P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model. *p<0.05, **p<0.01, ***p<0.001, ns p>0.05.

Zebrafish have two Atp1a3 paralogs, atp1a3a and atp1a3b. Similar to dmist, atp1a3a is widely expressed in the larval zebrafish brain (Figure 5F, compare to Figure 2D). While atp1a3b is also expressed in the zebrafish brain, its expression is more limited to regions of the midbrain and hindbrain (Figure S5C). To test whether these genes are involved in regulating zebrafish sleep, we used CRISPR/Cas9 to isolate an allele of atp1a3a containing a 19 bp deletion and an allele of atp1a3b containing a 14 bp deletion. Both mutations are predicted to generate null alleles due to deletion of the start codon (Figure 5G, S5D). Both atp1a3aΔ19/Δ19 and atp1a3bΔ14/Δ14 mutant larvae were healthy and viable through early development, although atp1a3b mutant larvae were not obtained at Mendelian ratios (55 wild type [52.5 expected], 142 [105] atp1a3b+/-, 13 [52.5] atp1a3b-/-; p<0.0001, Chi-squared), suggesting some impact on early stages of development leading to lethality. Contrary to a previous report based on morpholino injections (Doǧanli et al., 2013), neither mutant had defects in the inflation of their brain ventricles. Sleep-wake tracking experiments found that atp1a3bΔ14/Δ14 mutants were more active during the day with minimal sleep phenotypes (Figure S5E-G). In contrast, mutation of atp1a3a resulted in large effects on sleep-wake behaviour. Compared to wild type and heterozygous mutant siblings, atp1a3aΔ19/Δ19 animals were hyperactive throughout the day and night and had a large reduction in sleep at night (Figure 5H, I). The night-time sleep reduction was due to a reduction in the length of sleep bouts, as atp1a3a mutants even had a small increase in the number of sleep bouts at night (Figure 5J). In conclusion, loss of atp1a3a results in sleep loss at night, similar to treatment with the small molecule ouabain, and to dmist and fxyd1 mutants. Notably, the atp1a3a mutant phenotype is much stronger, as might be expected if Dmist plays a modulatory, and Atp1a3a a more central, role in Na+/K+ pump activity.

Dmist modulates Na+/K+ pump function and neuronal activity-induced sleep homeostasis

The similar night-time reduction in sleep in dmist and atp1a3a mutants, combined with the similarities between Dmist and Fxyd1, suggested that Dmist may regulate the Na+/K+ pump. We therefore exposed wild type and mutant larvae to pentylenetetrazol (PTZ), a GABA- receptor antagonist that leads to globally heightened neuronal activity and elevated intracellular sodium levels that must be renormalized by Na+/K+ pump activity. Consistent with the hypothesis that Dmist and Atp1a3a subunits are important for a fully functional Na+/K+ pump, brains from both dmisti8/i8 and atp1a3aΔ19/Δ19 larvae had elevated intracellular sodium levels after exposure to PTZ (Figure 6A). Thus, neither dmist nor atp1a3a mutants were able to restore intracellular sodium balance after sustained neuronal activity as quickly as wild type siblings. Consistent with the night-specific alterations in sleep behaviour, we also found that baseline brain Na+ levels in dmist mutants were significantly elevated at night but not during the day (Figure 6B). Collectively, these data are consistent with the hypothesis that night-time sleep duration is affected by changes in Na+/K+ pump function and that Dmist is required to maintain this function both at night and after sustained high levels of neuronal activity.

dmist mutants have altered sodium homeostasis

A) Brain sodium levels are significantly elevated after exposure to PTZ in both atp1a3aΔ19/Δ19 (2 independent experiments) and dmisti8/i8 (4 independent experiments) fish relative to wild type and heterozygous mutant siblings, as measured by fluorescence intensity of Sodium Green, normalized to the sample mean intensity. Crosses show mean ± SEM. n indicated the number of animals. Below are example images of brains stained with Sodium Green. *p<0.05, **p<0.01, one-way ANOVA, Tukey’s post hoc test.

B) Under baseline conditions, brain sodium levels are significantly elevated in dmisti8/i8 fish at night but not during the day, as measured by fluorescence intensity with Sodium Green. Crosses show mean ± SEM. *p<0.05, **p<0.01, one-way ANOVA, Tukey’s post hoc test.

C) dmisti8/i8 larvae have increased rebound sleep compared to wild type siblings following exposure to 5mM PTZ. Representative sleep traces of dmist+/+ (no drug, water vehicle controls in black; PTZ exposed in blue) and dmisti8/i8 (no drug in purple; PTZ exposed in red) following 1 hr exposure to 5 mM PTZ (black bar) in the morning. Data are mean ± SEM. dmisti8/+ animals are not plotted for clarity but are included in panel D.

D) Rebound sleep after exposure to 5 mM PTZ, calculated from the experiment in C. Each dot represents a single fish, grey lines show mean ± SEM.

E) Effect size of change in sleep after 1 hr treatment with 5 mM PTZ (and washout) compared to vehicle treated controls (error bars show 95% confidence intervals). *p<0.05, one-way ANOVA, Tukey’s post-hoc test.

F) Effect sizes (and 95% confidence interval) relative to wild types (dotted line) on sleep at night in larvae from dmist+/-; atp1a3a+/- in-crosses from 3 independent experiments. P-values are assigned by an F-test on the fixed effects coefficients from the linear mixed effects model relative to dmist+/+; atp1a3a+/+ animals. For all sleep-wake parameters, see Figure S6. *p<0.05,

**p<0.01, ***p<0.0001, ns p>0.05.

We have previously shown in zebrafish that a brief exposure to hyperactivity-inducing drugs such as the epileptogenic PTZ or wake-promoting caffeine induces a dose-dependent increase in homeostatic rebound sleep following drug washout that is phenotypically and mechanistically similar to rebound sleep following physical sleep deprivation (Reichert et al., 2019). Based on their exaggerated intracellular Na+ levels following exposure to PTZ, we predicted that dmist mutants would also have increased rebound sleep in response to heightened neuronal activity. Upon wash-on/wash-off of lower dose (5 mM) PTZ, sleep rebound occurs in approximately 50% of wild type larvae (Reichert et al., 2019; Figure 6C, D). In contrast, all dmisti8/i8 larvae showed increased rebound sleep compared to dmist+/+ sibling controls (Figure 6C-E). Taken together with the elevated sodium retention experiments, such increases in rebound sleep induced by neuronal activity suggests that dmisti8/i8 fish more rapidly accumulate sleep pressure in response to heightened neuronal activity.

Finally, we predicted that if Dmist is affecting baseline sleep via modulation of Atp1a3a- containing Na+/K+ pumps, dmist-/-; atp1a3a-/- double mutants should have a reduction in night- time sleep that is not the sum of effects from either mutant alone. In other words, if Dmist and Atp1a3a are acting in separate pathways, the double mutant would have an additive phenotype, but if Dmist and Atp1a3a act together in the same complex/pathway, the mutant phenotypes should be non-additive. Indeed, dmist-/-; atp1a3a-/- mutants have a sleep reduction similar to that of atp1a3a-/- mutants alone, consistent with a non-additive effect (Figure 6F and S6). Similar non-additivity can be also observed in the dmist-/-; atp1a3a+/- animals, which, like atp1a3a+/- animals alone, have a milder sleep reduction, indicating that the lack of additivity between dmist and atp1a3a phenotypes is unlikely due to a floor effect, since double homozygous mutants can sleep even less (Figure 6F). This genetic interaction data is consistent with our hypothesis that Atp1a3a and Dmist act in the same pathway—the Na+/K+ pump-- to influence sleep.

Discussion

Genetic screening discovers dmist, a novel sleep-regulatory gene

Using a reverse genetic viral screening strategy, we discovered a short-sleeping mutant, dmist, which has a disruption in a previously uncharacterized gene encoding a small transmembrane peptide. Given that the dmist mutant appeared within the limited number of 26 lines that we screened, it is likely that many other sleep genes are still waiting to be discovered in future screens. In zebrafish, one promising screening strategy will be to employ CRISPR/Cas9 genome editing to systematically target candidate genes. Advances in the efficiency of this technology now makes it feasible to perform a CRISPR “F0 screen” in which the consequences of bi-allelic, gene-specific mutations are rapidly tested in the first generation, with only the most promising lines pursued in germline-transmitted mutant lines (Grunwald et al., 2019; Jao et al., 2013; Kroll et al., 2021; Shah et al., 2015; Shankaran et al., 2017; Wu et al., 2018). CRISPR F0 screens could be scaled to systematically target the large number of candidate sleep-regulatory genes identified through human GWAS studies and sequencing of human patients suffering from insomnia and neuropsychiatric disorders (Allebrandt et al., 2013; Dashti et al., 2019; Jansen et al., 2019; Jones et al., 2019; Lane et al., 2019; Lek et al., 2016; Palagini et al., 2019).

Dmist is related to the Na+/K+ pump regulator Fxyd1

The small Dmist transmembrane protein is highly conserved across vertebrates, expressed in neurons, and important for maintaining normal sleep levels. How can such a small, single pass transmembrane protein lacking any clear functional domains modulate the function of neurons and ultimately animal behaviour? The recognition that Dmist has sequence homology (∼35% amino acid similarity; a conserved ‘RRR’ motif in the C-terminus) and structural homology (e.g. signal peptide and single pass transmembrane domains) to the Na+,K+-ATPase pump regulator Fxyd1 (Phospholemman) offers some important clues.

Fxyd1/Phospholemman is a member of the FXYD protein family, of which there are seven mammalian members (Sweadner and Rael, 2000). Each of the FXYD proteins is small, contains a characteristic FXYD domain, and has a single transmembrane domain. FXYD family members interact with alpha subunits of the Na+,K+ ATPase to regulate the function of this pump, with individual family members expressed in different tissues to modulate Na+,K+- ATPase activity depending on the physiological needs of the tissue (Geering et al., 2003). In cardiac muscle, FXYD1 is thought to act as a hub through which various signalling cascades, such as PKA, PKC, or nitric oxide, can activate or inhibit Na+ pump activity (Pavlovic et al., 2013). For example, FXYD1 is critical for mediating the increased Na+ pump activity observed after β-receptor stimulation via cAMP-PKA signalling (Despa et al., 2008). Much less is known about the role of FXYD1 in non-cardiac tissue, although it is expressed in neurons in the mammalian cerebellum, the choroid plexus, and ependymal cells, where it interacts with all three alpha subunits of the Na+,K+ ATPase (Feschenko et al., 2003).

In zebrafish, we also found that fxyd1 is expressed in cells around the ventricles and in the choroid plexus (Figure 4C), in contrast to dmist which is expressed in neurons throughout the brain. Despite the different expression patterns, mutation of each gene resulted in a similar reduction of sleep at night. However, unlike dmist mutants, which have fewer sleep bouts (i.e. initiate sleep less) and an increase in waking locomotor activity, fxyd1 mutants have shorter sleep bouts (i.e. cannot maintain sleep) on average and do not have a locomotor activity phenotype. Just as the various FXYD family members modulate the Na+/K+ pump in different tissue- and context-specific ways, this phenotypic variation between fxyd1 and dmist mutants could be due to the different fxyd1 and dmist expression patterns, modulation kinetics of pump/channel dynamics, or interaction with different accessory proteins or signal transduction cascades. Nevertheless, the similar timing and magnitude of sleep reduction, combined with the structural similarity of Fxyd1 and Dmist, suggest that they may regulate similar sleep- related processes.

Dmist, the sodium pump, and sleep

The similarity between Dmist and FXYD1 led us to directly manipulate the Na+,K+ ATPase to test its importance in sleep. The Na+,K+-ATPase is the major regulator of intracellular Na+ in all cells and, by actively exchanging two imported K+ ions for three exported Na+ ions, is essential for determining cellular resting membrane potential (reviewed in Clausen et al., 2017). The Na+,K+-ATPase consists of a catalytic alpha subunit (4 known isoforms, ATP1A1- 4), a supporting beta subunit (3 isoforms, ATP1B1-3), and a regulatory gamma subunit (the FXYD proteins). The alpha1 and alpha3 subunits are the predominant catalytic subunits in neurons (alpha2 is mostly restricted to glia), although the alpha1 subunit is also used ubiquitously in all tissues (McGrail et al., 1991). By mutating zebrafish orthologs of Atp1a3, we therefore could test the neuronal-specific role of the Na+,K+-ATPase in sleep.

Mutations in both zebrafish Atp1a3 orthologs increased waking locomotor behaviour during the day. However, only mutations in atp1a3a, which is expressed brain-wide, but not in atp1a3b, which is expressed in more restricted brain regions, led to changes in night-time sleep. The atp1a3a mutants have a larger sleep reduction than dmistvir, dmisti8, or fxyd1Δ28 mutants, which is expected since loss of a pump subunit should have a larger effect than the loss of a modulatory subunit, as has been shown for other ion channels (Cirelli et al., 2005; Wu et al., 2014). Autosomal dominant missense mutations leading to loss of function in ATP1A3 cause movement disorders such as rapid-onset dystonia parkinsonism and childhood alternating hemiplegia (recurrent paralysis on one side) in humans (Canfield et al., 2002; Heinzen et al., 2014), while loss of function mutations in Atp1a3 result in generalised seizures and locomotor abnormalities, including hyperactivity, in mice, which was not observed in zebrafish (Clapcote et al., 2009; Hunanyan et al., 2015; Ikeda et al., 2013; Kirshenbaum et al., 2011; Sugimoto et al., 2014). A very high prevalence of insomnia was recently reported in patients with childhood alternating hemiplegia, some of which harboured mutations in Atp1a3 (Kansagra et al., 2019), consistent with our observations that insomnia at night is a direct behavioural consequence of atp1a3a mutation in zebrafish. Since zebrafish atp1a3a mutants phenocopy the insomnia and hyperactivity phenotypes observed in patients, small molecule screens aimed at ameliorating zebrafish atp1a3a mutant phenotypes may be a promising approach for the rapid identification of new therapies for the management of this disease (Hoffman et al., 2016; Rihel et al., 2010).

Together, the night-specific sleep phenotypes of dmist, fxyd1, and atp1a3a mutants point to a role for the Na+,K+-ATPase in boosting sleep at night. How might the alpha3 catalytic subunit of the Na+/K+ pump regulate sleep, and how could Dmist be involved? We found that Dmist is required for proper maintenance of brain intracellular Na+ levels at night but not during the day, mirroring the timing of sleep disruption in dmisti8/i8 animals. This suggests that the decreased night-time sleep of dmist mutants is due to a specific requirement for Dmist modulation of the Na+/K+ pump at night. However, we cannot exclude the possibility that Dmist’s function is required in only a subset of critical sleep/wake regulatory neurons during the day that then influence behaviour at night, such as the wake-active, sleep-homeostatic regulating serotonergic neurons of the raphe (Oikonomou et al., 2019) or wake-promoting Hcrt/orexin neurons (Li et al., 2022). We also cannot exclude a role for Dmist and the Na+/K+ pump in developmental events that impact sleep, although our observation that ouabain treatment, which inhibits the pump acutely after early development is complete, also impacts sleep, argues against a developmental role. Another possibility is that disruption of proper establishment of the Na+ electrochemical gradient in dmist mutant neurons leads to dysfunction of various neurotransmitter reuptake transporters, including those for glycine, GABA, glutamate, serotonin, dopamine, and norepinephrine, which rely on energy from the Na+ gradient to function (Kristensen et al., 2011).

A third possibility is that Dmist and the Na+,K+-ATPase regulate sleep not by modulation of neuronal activity per se but rather via modulation of extracellular ion concentrations. Recent work has demonstrated that interstitial ions fluctuate across the sleep/wake cycle in mice. For example, extracellular K+ is high during wakefulness, and cerebrospinal fluid containing the ion concentrations found during wakefulness directly applied to the brain can locally shift neuronal activity into wake-like states (Ding et al., 2016). Given that the Na+,K+-ATPase actively exchanges Na+ ions for K+, the high intracellular Na+ levels we observe in atp1a3a and dmist mutants is likely accompanied by high extracellular K+. Although we can only speculate at this time, a model in which extracellular ions that accumulate during wakefulness and then directly signal onto sleep-regulatory neurons could provide a direct link between Na+,K+ ATPase activity, neuronal firing, and sleep homeostasis. Such a model could also explain why disruption of fxyd1 in non-neuronal cells also leads to a reduction in night-time sleep.

In addition to decreased night-time sleep, we also observed that dmist mutants have an exaggerated sleep rebound response following the high, widespread neuronal activity induced by the GABA-receptor antagonist, PTZ. Since both Atp1a3a and Dmist were essential for re- establishing proper brain intracellular Na+ levels following PTZ exposure (Figure 6A), we speculate that the exaggerated sleep rebound is a consequence of increased neuronal depolarization due to defective Na+ pump activity. This is consistent with our previous observations that the intensity of brain-wide neuronal activity impacts the magnitude of subsequent sleep rebound via engagement of the Galanin sleep-homeostatic output arm (Reichert et al., 2019). Why does loss of dmist lead to both decreased night-time sleep and increased sleep rebound in response to exaggerated neuronal activity during the day? One possibility is that Na+/K+ pump complexes made up of different alpha and beta subunits may be differentially required for maintaining Na+ homeostasis under physiological conditions and have different affinities for (or regulation by) Dmist. For example, the Atp1a1 subunit is considered the Na+/K+ pump workhorse in neurons, while Atp1a3, which has a lower affinity for Na+ ions, plays an essential role in repolarizing neurons when Na+ rapidly increases during high levels of neuronal activity, such as after a seizure (Azarias et al., 2013). If Dmist preferentially interacts with Atp1a3a subunit, with which the non-additive effect of dmist and atp1a3a mutation on sleep is consistent, day-time sleep-related phenotypes in dmist mutants might be uncovered only during physiological challenge. Conversely, neurons may be more dependent on Atp1a3a and Dmist for sodium homeostasis at night due to changes in Na+/K+ pump composition, Dmist interactions, or ion binding affinities. For example, activity of the Na+/K+ pump can be modulated by the circadian clock (Damulewicz et al., 2013; Nakashima et al., 2018), changes in substrate availability, including ATP (reviewed in Therien and Blostein, 2000), or hormones (Ewart and Klip, 1995). Teasing out how Dmist modulation of the Na+/K+ pump changes across the day-night cycle, and in which neurons Dmist’s function may be particularly important at night, will require future investigation.

In conclusion, through a genetic screening strategy in zebrafish, we have identified a novel brain expressed gene that encodes a small transmembrane protein regulator of night-time sleep and wake behaviours. Future work will be required to uncover the precise signalling dynamics by which Dmist regulates the Na+,K+-ATPase and sleep.

Supporting information